---
title: "Counting to 90 in Greenlandic and Faroese"
subtitle: Analysing the North Atlantic MPs Voting Behaviour in the Danish Parliament
authors:
  - name: Moritz Twente
    affiliation: University of Basel
    email: moritz.twente@unibas.ch
    orcid: 0009-0005-7187-9774
abstract: > 
    nordatlantisk-ft is compiled from data available on Folketingets Open Data Platform [@folketingetFolketingetsAbneData2026] and comprises 40899 records of votes cast by Northatlantic MPs as per the most recent update (February 06, 2026). The dataset documents voting behavior of MPs from Greenland and the Faroe Islands and also contains metadata on each record – e.g. date and topic of each vote or the overall voting result, covering votes on 10227 proposals in Folketinget. Refer to the [documentation](https://mtwente.github.io/nordatlantisk-ft) for details^[Both source code and data used for this paper are available on [GitHub](https://www.github.com/mtwente/counting-to-90).].
date: last-modified
date-format: long
bibliography: ../assets/nordatlantisk-ft.bib
#csl: ../assets/american-political-science-association.csl
reference-section-title: References
link-citations: true
lang: en
output-file: paper.pdf
filters:
  - wordcount
wordcount:
  count-code-blocks: false
  count-inline-code: false
format: 
  pdf: 
    documentclass: article
    classoption: twocolumn
    mainfont: Jost
    filters: [../assets/quotation-marks.lua]
    quot-lang: de-CH
    colorlinks: true
    toc: true
    toc-depth: 2
    geometry:
      - top=22mm
      - bottom=18mm
    include-in-header:
      text: |
        \usepackage{float}
        \usepackage{fancyhdr}
        \pagestyle{fancy}
        \fancyhf{} % clear all header and footer fields

        % first page footer
        \fancypagestyle{plain}{
        \fancyhf{}
        \fancyfoot[L]{\footnotesize Moritz Twente, 19-911-940, moritz.twente@unibas.ch}
        \fancyfoot[R]{\footnotesize \thepage}
        \renewcommand{\headrulewidth}{0pt}
        }

        % Right-justified header (smaller font)
        \fancyfoot[L]{\footnotesize Moritz Twente, 19-911-940, moritz.twente@unibas.ch}

        % Page number in footer (centered)
        \fancyfoot[R]{\footnotesize \thepage}

        % Remove header line (optional, cleaner look)
        \renewcommand{\headrulewidth}{0pt}

        % reduce whitespace below header
        \setlength{\headsep}{9pt}
---

<!-- 8000 -- 10'000 words, 20--25 pages excl. Bibliography -->

<!-- writing: 
- every paragraph includes one argument/thought
- structure of a paragraph: topic sentence, development of the argument (2-3 sentences that elaborate your idea, e.g. by example)
- concluding sentence, link to the next paragraph -->

```{r setup}
#| echo: false
#| message: false

# Setup -----
## Packages -----

library(here)
library(ggplot2)
library(dplyr)
library(readr)
library(knitr)
library(forcats)
library(lubridate)
library(scales)
library(showtext)
library(xfun)
library(patchwork)
suppressMessages(library(stargazer))
library(kableExtra)

## Install Font -----

font_add("Jost",
  regular = here("assets", "fonts", "ttf", "Jost-400-Book.ttf"))

showtext_auto()

## Read Data -----

### List of MPs -----
mp_names <- readRDS(here("data", "processed", "MP_names.rds"))

### List of political parties -----
political_parties <- readRDS(here("data", "processed", "political_parties.rds"))

### Voting Records Database -----
northatlantic_ft <- readRDS(here("data", "processed", "northatlantic_ft.rds"))

### Roll Call Database -----
roll_call_results_ft <- readRDS(here("data", "processed", "roll_call_results_ft.rds"))

### Timeline of Mandates -----
MP_dates <- readRDS(here("data", "processed", "MP_dates.rds"))
elections <- readRDS(here("data", "processed", "election_dates.rds"))

#### Convert dates
MP_dates <- MP_dates %>%
  mutate(
    start = ymd(start),
    end   = ymd(end),
    # Replace NA in "end" with today's date
    end   = if_else(is.na(end), Sys.Date(), end)
  )

## Manipulate Data -----

### Avoid confusing MPs with the same last name
MP_dates <- MP_dates %>%
  mutate(
    plotting_name = case_when(
      MP_id == "12283" ~ "A. V. Johannesen",
      MP_id == "6687"  ~ "K. L. Johannesen",
      TRUE             ~ surname
    )
  )

northatlantic_ft <- northatlantic_ft %>%
  mutate(
    plotting_name = case_when(
      MP_id == "12283" ~ "A. V. Johannesen",
      MP_id == "6687"  ~ "K. L. Johannesen",
      TRUE             ~ surname
    )
  )

### Order origin factor levels for facet order
northatlantic_ft <- northatlantic_ft %>%
  mutate(origin = forcats::fct_rev(origin))

### Order dataset by MP surnames for plotting
northatlantic_ft_ordered <- northatlantic_ft %>%
  mutate(surname = fct_reorder(surname, vote_type_id, .fun = length)) %>%
  mutate(plotting_name = fct_reorder(plotting_name, vote_type_id, .fun = length)) %>%
  mutate(vote_type_id = fct_relevel(vote_type_id, "1", "2", "4", "3"))

### Subset only with votes that were actually cast
northatlantic_ft_without_absences <- subset(northatlantic_ft_ordered, vote_type_id != 3) %>%
  mutate(surname = fct_reorder(surname, vote_type_id, .fun = length)) %>%
  mutate(plotting_name = fct_reorder(plotting_name, vote_type_id, .fun = length))

### Subset Election Dates for Folketinget only

elections_ft <- elections %>%
  filter(election_type == "Folketing Election")

## Set Plotting Theme -----

theme_nordatlantisk_basis <- function(){
  
  theme_grey(base_family = "Jost") %+replace%
    theme(axis.title = element_blank(),
          axis.ticks.y = element_blank(),
          panel.background = element_blank(),
          panel.grid.major.x = element_line(color = "#000000", linewidth = 0.1),
          panel.grid.minor.x = element_line(color = "#000000", linewidth = 0.1),
          panel.grid.major.y = element_blank(),
          panel.grid.minor.y = element_blank(),
          axis.ticks.x = element_line(color = "#000000", linewidth = 0.1),
          axis.line.y.left = element_line(color = "#000000", linewidth = 0.1))
}

```

# Counting to 90

<!-- into according to powner: three-paragraph framework for a 20p paper -->

<!-- 1. establish puzzling empirical facts, articulate contradiction/puzzle, briefly describe how others have attempted to solve it in the past -->

Four out of the 179 MPs in the parliament of Denmark, Folketinget, represent Greenland and the Faroe Islands (referred to collectively as Northatlantic MPs). They usually only position themselves on issues where their constituencies, as parts of the Danish Realm, cannot decide independently[^issues]. Party representatives state that they do not want to interfere in 'domestic' Danish politics [@kurrild-klitgaardElectionInversionsCoalitions2013, 130] and thus generally refrain from taking part in parliamentary work outside those policy areas. This setup has been widely accepted for decades, going so far that those MPs, i.e. elected members of a legislative assembly, are labelled as "non-political" [@millerFriendsRivals1996, 72] politicians[^slesvig]. But in times of less stable majorities, when one of the two Danish party blocs finds themselves without a 'domestic majority' (*indenrigspolitisk flertal*), Northatlantic MPs may flip the outcome of a roll call vote in Folketinget or prove crucial for securing the coalition's stability [@harderFaeroskeOgGronlandske2022, 19] by helping to form a majority of 90 seats -- to use the Danish expression: to be able to count to 90.

[^issues]: This pertains to issues about e.g. foreign affairs, defence and security, currency and financial regulations, citizenship, the police and judiciary systems, aviation regulations etc. [@cleggParliamentaryRepresentationOverseas2022, 235].

[^slesvig]: Miller, referring to the results of the 1957 election, also includes a representative for the German-speaking minority in Southern Slesvig in this group of non-political MPs. However, in this paper I will concentrate on Greenlandic and Faroese MPs only, not considering the representation of other minorities in Folketinget.

These situations regularly cause controversy among politicians from 'continental' Denmark, focussing on the symbolic aspects of the Northatlantic presence in Folketinget, not least in the context of post-colonialist debates on the relations between Greenland, the Faroe Islands and Denmark -- leading to heightened tensions and even bullying [@harderArcticIndigenousRepresentation2024, 141; @cleggParliamentaryRepresentationOverseas2022, 239] with Danish politicians employing colonialist metaphors [e.g. 'buying a Greenlander', cf. @kamstrupNuHarVi2017; @larsenSkatteministerTilNytarskur2017; @sjolieDanmarkManglerOpgor2008]. Additionally, the Northatlantic MPs' positions appear to be a fixture of news coverage leading up to elections, sometimes coming across as a quirk of Denmark's political system [e.g. @dallGronlandOgFaeroerne2022; @engelundMorgenpostFraGronland2026a; @prakashBlaEllerRod2026].  Overall, Greenlandic and Faroese representation in Folketinget is brought up when deemed convenient [@harderFaeroskeOgGronlandske2022, 5; @veirumForskerNordatlantiskIndflydelse2022].

The parliamentary work of the Northatlantic MPs can still be considered a black box, and is only starting to become a topic of scientific investigation. So far, a summary of research on this topic is: If they find themselves in situations where they have power, they tend to use it [@harderFaeroskeOgGronlandske2022, 18; @skjaevelandModellingGovernmentFormation2009, 723]. Recently, parliamentary questions and committee work have been the vantage points from which the Northatlantic MPs's work is analysed [@westDanishRealmArctic2025; @harderFaeroskeOgGronlandske2022; @harderCaseInnovativeParliamentary2022], as has been cross-parliamentary control [@harderCrossparliamentaryControl2025]. Their influence on coalition formation is another angle, but coalition research rather focuses on Danish (median) parties' influence [cf. @skjaevelandModellingGovernmentFormation2009]. Overall, however, we still know little about the activities of Greenlandic and Faroese MPs [@harderFaeroskeOgGronlandske2022, 5], especially compared to the amount of anecdotic evidence and news coverage available over the years.

>> but why should we care?

<!-- 2. focus primarily on theory and findings of my work, show how it significantly improves over previous work -->

Individual voting behavior is a 

but is there more to it?

they are usually not official members of cabinet, except for Mikael Gam [@skjaevelandMindretalsregeringerDanmarkFor2003]

"Moving to substantive representation, studies of the Danish and Canadian parliaments suggest that Indigenous representatives are especially engaged in acting for Indigenous interests." [@harderArcticIndigenousRepresentation2024, 143]

DK using GL/FO (and their FT membership) to legitimize themselves as Arctic state [@westDanishRealmArctic2025, 589]

<!-- 3. roadmap of the rest of the paper, per paragraph/section style -->




- inner-Rigsfællesskabet debates very current,
  - mechanisms of Rigsfællesskabet dynamics, but within Folketinget?
- opens up international postcolonialism debates

## Research Questions, Relevant X/Y variables

1. In general, when do the Northatlantic MPs take part in roll call votes?
2. The 2x2 MPs are conceived as one group by DK politicians and media. Do they also vote like one group?
3. When Northatlantic MPs have the chance to have influence, they act on it (based on decades of more or less anecdotal evidence). So, does the seat share of Danish coalition parties influence voting behaviour of Northatlantic MPs?
4. can the Northatlantic MP's voting behavior also be considered a form of cross-parliamentary control?

* In general, when do the Northatlantic MPs take part in roll call votes? Hypothesis: The MPs are more likely to take part in voting when the issue concerns Greenland/policy areas that GL/FO do not have authority over (e.g. Defense, based on keywords in proposal description and the ministerområde responsible for the proposal, e.g. Forsvarsministeriet, Finansministeriet etc.)
* The 2x2 MPs are conceived as one group by DK politicians and media. Do they also vote like one group? Hypothesis: After excluding absences, similar voting behavior can not be observed across Greenlandic and Faroese MPs, but across the political blocks the MPs belong to (left/right -- in the dataset: usually 2x left from GL, and 1x each from FO, but never multiple MPs from the same party at the same time)
* When Northatlantic MPs have the chance to have influence, they act on it (based on decades of more or less anecdotal evidence). So, does the seat share of Danish coalition parties influence voting behaviour of Northatlantic MPs? Hypothesis: The fewer FT mandates the Danish cabinet coalition parties have themselves, the more likely are GL/FO MPs to cast a vote (regardless of the content of the proposal being voted on)
  * Based on questioning/committee activity analysis of Northatlantic MPs: Cross-parliamentary control mechanisms are especially made use of if the MP's own party is an opposition party in GL/FO. Transferring this finding onto another type of parliamentary activity, can the Northatlantic MP's voting behavior also be considered a form of cross-parliamentary control? Hypothesis: MPs are more likely to engage in parliamentary activity in FT, in this case taking part in voting, if their own party is in opposition in GL/FO.

# Literature Review

<!-- - summary of relevant literature for your research question!
- aim of literature review: identify research gap (and by this your own contribution)
- do not mention every paper written about your case but the most important+related ones
- pathbreaking papers and latest developments should be included
- Direct citations only if absolutely needed, better are paraphrasing -->

Over the past decades, the activity of Folketing MPs from Greenland and the Faroe Islands has been written about time and again -- usually when there are narrow majorities and the 'Northatlantic' votes might become relevant to the voting outcomes, e.g. after elections in 2022, 2015, 2011 or 2007 [@harderFaeroskeOgGronlandske2022; @sorensenEfterBrokBla2022]. While situations like these are regularly reported on by journalists [e.g. @nathanGronlandskMferSer2024; @veirumPartilederePatalerGronlandsk2022; @frostLokkeRaserNordatlantiske2022; @skouKaareFireMandater2008][^zotero], the attention they receive from researchers has long been marginal.

[^zotero]: For a larger collection of reports in mostly Danish-language media from recent years, see my [nordatlantisk-ft Zotero collection](https://www.zotero.org/groups/5346749/nordatlantisk-ft).

If Greenlandic or Faroese political representation in Folketinget does appear in research output, it is often referred to in anecdotes [e.g. "A coalition saved by Greenland", @millerFriendsRivals1996, 80] or in parentheses [e.g. @klintDanishLegislatorsDatabase2023; @elklitElectoralSystemFair2020; @nannestadPolitischeSystemDanemarks2009; @kjaerDanskeFolketingsmedlemmerParlamentarisk2004], acknowledging its existence, but mostly, sometimes explicitly, not considering it for analysis [e.g. @kuselerDetectingClimateDelay2026; @ejnarhansenReconsideringPartyDistances2008; @skjaevelandPartyCohesionDanish2001; @valenGeographyPoliticalRepresentation2000; @westPoliticalPartiesParliamentary2025; @elklitDenmarkSimplicityEmbedded2005; @christiansenMinorityCoalitionGovernance2014]. Given the anecdotal evidence about the Northatlantic MPs' presence and the often heated reactions to situations where roll call results are depending on votes from (at least one of) them, it is striking how research related to Folketinget and its dynamics has largely not taken up on this aspect. In the following paragraphs, I will take a closer look at literature on the functioning of Folketing itself, on coalition and government formation, on the Danish election system as well as on the political parties.

Writing about parliamentary cross-bloc cooperation, @green-pedersenBlokpolitikOgNye2020 [311] do not include political parties from Greenland and the Faroe Islands when counting the number of parties that are represented in parliament at all. Neither does @helboepedersenParliamentFOLKETINGETPowerful2020 in her investigation into the institutional powers of Folketinget, although she briefly mentions the existence of Northatlantic MPs. @hansenCommitteeAssignmentPolitics2010 states he is not considering these MPs in an analysis of committee assignments. In a categorisation of roll call votes, @skjaevelandPartyCohesionDanish2001 explicitly excludes matters concerning Northatlantic MPs from a category of domestic issues.

With regard to the composition of Folketinget, @kurrild-klitgaardElectionInversionsCoalitions2013 analyses election inversions, meaning the party/political bloc winning the most votes does not end up with the most parliamentary seats after the election -- a situation the Northatlantic MPs have ended up in repeatedly. Using the 1971 election, he demonstrates that the lower cost of winning a Folketinget mandate in the Faroe Islands resulted in a margin so narrow that, if added to continental Danish overall results, would have resulted in an alternative majority [-@kurrild-klitgaardElectionInversionsCoalitions2013, 131]. This is, however, not considered to be a flaw, and the election system of Denmark is regarded to be exceptionally well-functioning [@elklitElectoralSystemFair2020; @lijphartForeword2005], even though that assessment, again, excludes elections in Greenland and the Faroe Islands [cf. @elklitDenmarkSimplicityEmbedded2005]. An otherwise comprehensive database with biographic information on Folketinget legislators by @klintDanishLegislatorsDatabase2023 also does not include anyone from Greenland and the Faroe Islands.

The Northatlantic MPs are sometimes factored in as relevant actors in publications on coalition/government formation in Denmark, especially by @skjaevelandGovernmentFormationDenmark2003[^fernleihe]. He touches upon the Northatlantic MPs' importance in his research on government formation in Denmark [-@skjaevelandModellingGovernmentFormation2009], acknowledging their importance in some coalition life cycles. The take-away from his section on the North Atlantic Folketing is that "the more important their votes, the more involved they have become" [-@skjaevelandModellingGovernmentFormation2009, 723]. But even when acknowledging the voting behavior of North Atlantic MPs as a potentially decisive factor for government formations, they are still only partially relevant to research on the distribution of power in Danish party politics, which tends to focus on median parties such as Radikale Venstre [@green-pedersenBlocPoliticsVs2005; @green-pedersenDetRadikaleVenstres2003; @skjaevelandMindretalsregeringerDanmarkFor2003].

[^fernleihe]: @harderFaeroskeOgGronlandske2022 [7] characterises Skjæveland's PhD thesis as a comprehensive coverage of the Northatlantic MPs influence on coalition formations in Denmark. But, as accessing this publication requires more effort than is reasonable for a seminar paper like this, I refer to @skjaevelandModellingGovernmentFormation2009 instead, which builds on the PhD thesis published in 2003.

Research on political parties in Denmark, e.g. by @green-pedersenPartySystemOpen2020 or @pedersenPartyDistancesDanish1971, does not cover parties from Greenland or the Faroe Islands. This is hardly surprising, given that the political parties competing in elections in Denmark, Greenland and the Faroe Islands are separate entities and each party can only be voted for from within either of the three parts of the Danish Realm [@kurrild-klitgaardElectionInversionsCoalitions2013]. Likewise, research on political parties of Greenland and the Faroe Islands features a domestic focus as well [e.g. @ahlnessNunattaQitornaiParty2020; @westSkipanarligarFortreytirOg2022; @michelsenGronlandPartisystemUdvikling1979]. Institutional associations between Greenlandic and Danish parties linked to parliamentary work in Folketinget are mentioned by @ackrenPoliticalPartiesGreenland2015, building on ideologically similar backgrounds of Siumut of Greenland and Socialdemokratiet in Denmark resp. of Atassut (GL) and Venstre (DK). However, Folketinget representation does not seem to be a central issue for regional parties in Greenland and the Faroe Islands. Because Greenland and the Faroe Islands only hold two seats each, @ackrenExploringRoleRegional2019 deem it unlikely that any Northatlantic proposal would gain enough traction in Folketinget anyway. For these regional parties, the preferred mode of conducting relations with Denmark is by communicating between government officials instead [@ackrenExploringRoleRegional2019; @harderArcticIndigenousRepresentation2024, 130].

The inclusion of anything related to the Greenlandic and Faroese representation within the Danish Realm into analyses of the political system of Denmark is not trivial, not least due to the separate party systems and because there are no institutional links between the Northatlantic MPs and neither their national governments nor parliaments [@westDanishRealmArctic2025, 602][^doublemandate]. It is likely in this context that the activities of the four Northatlantic MPs feature only in the margins of research on Danish politics. This can be said for research on Folketing and its inner workings specifically, but also for adjacent topics such as research on elections and coalition formations as summarised above.

[^doublemandate]: Greenlandic and Faroese representatives can be MPs in their local parliaments at the same time as they are members of Folketinget. For example, Folketinget member Aki-Matilda Høegh-Dam was elected in the 2025 Inatisartut election, thus serving as MP in the Danish and Greenlandic parliaments at the same time (until the 2026 Folketinget election). In the Faroe Islands, this type of double mandates has been "heavily criticized" [@westDanishRealmArctic2025, 602]. 

With this in mind, I assume that the activities of the Greenlandic and Faroese members of Folketinget have mostly been regarded not as a relevant aspect of the political system of Denmark that is worth considering. In conclusion, there has been very little research activity on MPs from Greenland and the Faroe Islands and their behavior in Folketinget compared to the importance that is repeatedly being attributed to their presence [cf. @harderFaeroskeOgGronlandske2022, 5]. This has, however, changed in recent years with publications explicitly targeting Northatlantic representation in Folketinget which I am going to focus on in the following paragraphs.

<!-- Section Voting Behavior -->
@harderFaeroskeOgGronlandske2022 investigates the North Atlantic MPs' behavior between 2005 and 2019 by looking at parliamentary questions posed by them as well as their participation in committee work. In her paper, the first to systematically research the activities of the Greenlandic/Faroese Folketing representations, she finds that for both questions and committee work, Greenlandic MPs have been more active than their Faroese counterparts. Overall, they are most active around topics with direct relevance for their constituencies, and some MPs around issues that concern Greenlandic and Faroese inhabitants of Denmark proper as well [cf. @cleggParliamentaryRepresentationOverseas2022, 238]. Long-term strategies regarding the relationship between the members of the Danish Realm do not appear to be a priority [@harderFaeroskeOgGronlandske2022, 15]. Harder's analysis suggests that there is no correlation between the number of questions asked and whether the MP's party is part of the cabinet in Greenland or the Faroe Islands. She concludes that MPs tend to support the political bloc that aligns most closely with their own party's positions. On the topic of securing majorities in roll call votes, she echoes findings for the second half of the 20^th^ century by @skjaevelandGovernmentFormationDenmark2003: MPs take part in roll call votes in particular if there is a chance their vote flips the outcome -- which usually leads to debates about the North Atlantic's influence in 'domestic' politics.

<!-- Section (Cross-)Parliamentary Control -->
Apart from their voting behavior, Northatlantic MPs are also investigated in the context of parliamentary control mechanisms. While Folketinget as an institution is equipped with relatively strong mechanisms for controlling the (Danish) government [@helboepedersenParliamentFOLKETINGETPowerful2020, 102], it is another question whether control is exercised vis-à-vis Naalakkersuisut of Greenland and Landsstýrið of the Faroe Islands as well. @harderCaseInnovativeParliamentary2022 use parliamentary questions as case study for investigating cross-parliamentary control -- if/how Northatlantic MPs use their mandates in Folketinget to check on the (in)actions of their local governments. This can be said to be the case for 7.8% of the questions posed to Danish ministers between 2005 and 2020. Harder and West find that especially MPs from parties known to be sceptical of the Danish Realm are making use of this form of cross-parliamentary control, from Faroese parties in particular. Greenlandic MPs are found to address a broader range of issues compared to the ones from the Faroe Islands [@westDanishRealmArctic2025, 598]. Harder then elaborates on prerequisites and effects of cross-parliamentary control mechanisms in a second-order parliament -- which Folketinget is from a Greenlandic/Faroese perspective -- hypothesising that MPs "will especially engage in cross-parliamentary control when they belong to parties that are in opposition within the government system where they seek to control the government" [-@harderCrossparliamentaryControl2025, 12], bearing in mind that in her dataset, this observation is based of the behavior of representatives of one political party only. Additionally, party group membership appears to be a variable of importance: All but one of the observed cases of cross-parliamentary control occur after the disbandment of the Northatlantic Parliamentary Group in 2011, indicating that party competition is a prerequisite.

In addition to posing questions, committee membership is an effective tool of parliamentary control in Folketinget [@holliDualApproachComparing2016]. Since the inception of the current committees setup in 1971, Greenlandic MPs have been members of parliamentary committees more often than their Faroese counterparts [@harderFaeroskeOgGronlandske2022, 11]. Northatlantic MPs are usually members of the committees on Greenland and the Faroe Islands. Other than that, they are active in committees concerning policy areas, especially foreign affairs, that are not in the responsibility of the local governments. Harder mentions Edmund Joensen and Aleqa Hammond as MPs who were members of comparably many committees and -- alluding to their importance in securing majorities -- suggests that their willingness to support the Danish government in issues that did not concern their constitutencies directly was met with the allocation of more committee seats than usual [@harderFaeroskeOgGronlandske2022, 17].

<!-- Section Rigsfællesskabet -->
This recent increase in research output regarding the Northatlantic representation in Folketinget has to be seen in the context of the changed status of the Arctic within the Danish Realm and beyond, leading to a pressure to open the "'black box' of the Danish state" [@westDanishRealmArctic2025, 589; cf. @harderFaeroskeOgGronlandske2022, 6] that the association of Denmark, Greenland and the Faroe Islands as *Rigsfællesskabet* and its dynamics were for a long time [cf. @hviidDetArktiskeVindue2022]. Denmarks "Arctic awakening" [@jacobsenDanmarksArktiskeUdenrigspolitik2021] in the beginning of the 21^st^ century led to a shift from understanding Danish-Greenlandic-Faroese relations as domestic affairs within the Danish Realm to focussing on geopolitics of the Arctic as a whole. The four Northatlantic MPs take center stage in that regard, as Folketinget allocates parliamentary positions that are central to Arctic policy to Northatlantic MPs [@westDanishRealmArctic2025, 603], not least in order to legitimize Denmark as a geopolitical player in the Arctic. Still, their presence has a considerable symbolic value, as noted by @cleggParliamentaryRepresentationOverseas2022 [237]. By highlighting diverging opinions as to whether the Northatlantic MPs also represent Greenlandic/Faroese inhabitants of Denmark[^representing], the authors bring up the perspective of overseas territories' representation.

[^representing]: In the 2026 election campaign, a Danish-Greenlandic candidate in Denmark campaigned as representing Greenlanders in Denmark, pledging to bring the number of 'Greenlandic' mandates up to 2.5 [@mollerMangearigHandboldspillerVil2026].

In a similar context, the new geopolitical status of the Greenlandic and Faroese parts of *Rigsfællesskabet* is accompanied by an increased awareness of the colonialist heritage of Denmark in both Greenland and, to a lesser extent, the Faroe Islands [cf. @jensenPostcolonialDenmarkRot2015; @dyrendomgraugaardColonialReproductiveCoercion2025; @kociGreenlandFaroeIslands2021]. However, national parliaments are not in the spotlight of research dealing with Arctic Indigenous representation [@harderArcticIndigenousRepresentation2024, 130] which, in this case, would mainly concern MPs from Greenland, and not the Faroe Islands anyway.

## Research Gap

As demonstrated above, the Greenlandic and Faroese representation in Folketinget has only recently become the target of scientific investigation. Because research on the parliament of Denmark that does not use a decidedly Arctic lens does not cover the Northatlantic MPs by and large, there are still a lot of unknowns about the activities of these politicians. Anecdotal evidence, also in the form of news reports, points to repeated situations where the four Northatlantic MPs of Folketinget do have influence on 'domestic' Danish lawmaking. My aim is to go beyond treating the parliamentary activity of Greenlandic and Faroese MPs as curiosity of the political system of Denmark and to figure out whether there is evidence that the influence of Northatlantic MPs is more than a move in the blame game of somebody who just lost a vote but instead a recurring issue.

In doing so, I am building on central papers mentioned above [esp. @harderFaeroskeOgGronlandske2022; @harderCrossparliamentaryControl2025; @westPoliticalPartiesParliamentary2025] and try to add to their findings in several ways: While published recently, they do not use data from later than 2020. In light of the political developments that have affected the equilibrium of *Rigsfællesskabet* in recent years -- from postcolonialist perspectives to international security and, most recently, U.S.-Denmark relations, to name a few -- it is warranted to take closer look at the status quo again, using the most recent data available. Additionally, I am looking at the Northatlantic MPs from a different angle than has been done before: While investigating MPs' activities by looking at committee activities and questions is a common and reasonable choice for unit of analysis [@cleggParliamentaryRepresentationOverseas2022, 238], I am focusing on individual voting records from roll call votes in plenary sessions.

Additionally, the topic at hand also touches upon other issues such as ethnic representation in parliaments as well as regional parties. Results from using the novel dataset used here also extend knowledge on these research areas, even though they were not explicitly part of the initial concept of this paper.

<!-- powner: how does my work fit into the existing literature? what's the current explanation for my dependent variable? -->

# Research Design

## Theoretical Approach

<!-- - case selection/overview data (temporal and spatial), unit of analysis
- measurement dependent variable (and data source)
- measurement independent and control variables (and data source)
- methodological approach (why this one?) -->

  - if they have power, they use it [@harderFaeroskeOgGronlandske2022, 18; @skjaevelandModellingGovernmentFormation2009, 723]
  + relevans efter valg 1971 [@gadGronlandskeFolketingsmedlemmer2022; @kurrild-klitgaardElectionInversionsCoalitions2013, 131; @skjaevelandModellingGovernmentFormation2009, 723], 2007 om asylcentre, but was resolved without GL/FO influence [@gadGronlandskeFolketingsmedlemmer2022; @cleggParliamentaryRepresentationOverseas2022, 239], relevans efter valg 2007, 2011, 2015 [@harderFaeroskeOgGronlandske2022, 19] and 1998 [@skjaevelandModellingGovernmentFormation2009, 723], and 1960 [@skjaevelandGovernmentFormationDenmark2003, 448], and 1957 [@millerFriendsRivals1996, 71]. 2015 not relevant after all? [@arndtFolketingswahlDanemarkVom2019, 785]
  - "I to tilfælde har grønlandsk valgte MFere opnået ministerposten" [@michelsenGronlandPartisystemUdvikling1979, 55] > Gam og hvem?
  - @dallGronlandOgFaeroerne2022 with comprehensive overview

  - "Den anden er Edmund Joensen (SP), der i perioden lige efter valget i 2007 spillede en afgørende rolle i at sikre den borgerlige regerings parlamentariske flertal. Som det eneste færøske medlem i hele perioden blev han fuldt medlem af hele fem folketingsudvalg samt suppleant i et enkelt. Noget tyder således på, at det at sikre regeringens flertal kan blive afspejlet i blandt andet antallet af udvalgsposter, som man råder over – en antagelse, der i øvrigt underbygges af, at Aleqa Hammond (SIU/NQ) blev tildelt den symbolsk vigtige rolle som formand for Grønlandsudvalget efter at have indgået aftale om at støtte den borgerlige regering" [@harderFaeroskeOgGronlandske2022, 17]
  - "Analysen ovenfor viser imidlertid, at der ikke er nogen signifikant sammenhæng mellem antallet af parlamentariske spørgsmål, som et nordatlantisk medlem stiller, og spørgsmålet om, hvorvidt vedkommendes partifamilie støttede regering eller opposition" [@harderFaeroskeOgGronlandske2022, 17]

"Overordnet set kan det fastslås, at de grønlandske medlemmer nu som før del-
tager langt mere aktivt i folketingsarbejdet, end de færøske medlemmer gør" [@harderFaeroskeOgGronlandske2022, 19]

"... parliamentary questions are typically considered the most relevant tool for oversight activity (Rasch, 1994; Wiberg, 1994). Yet questioning activity can be used for other purposes in addition to oversight and primarily is an activity for opposition party" [@westPoliticalPartiesParliamentary2025, 969]

"the assumption is that government party MPs lack oversight-related incentives, while the opposition party MPs have incentives to engage, but primarily in blame games (Weaver, 1986), to damage government reputation and inflict cost (Maor, 1999; Mayhew, 1974; West, 2018, 2020). In other words, parliamentary oversight is most likely realised as a by-product of oppositional partisan-motivated activity" [@westPoliticalPartiesParliamentary2025, 970]

## Hypotheses

<!-- * Hypothesis: causal relationship between two variables: the more, the less...
* Hypothesis in its own paragraph and in italics -->

## Empirical Strategy

<!-- - variables and controls
- unit of analysis
- method (logit model etc.) -->

see method used for Bundestag in @maiLoyalActivistsParty2024

"Fordi flere af de nordatlantiske medlemmer holder længere orlov fra Folketinget, er data herfor indhentet på parti- og ikke individniveau" [@harderFaeroskeOgGronlandske2022, 9]

"... it is custom in Danish legislative studies to use parliamentarians’ membership of legislative committees as an indicator of their political activities" [@cleggParliamentaryRepresentationOverseas2022, 238]

- distribution of ft_process_step? (or later during empirical strategy)
  - use tredje behandling? cf. @green-pedersenBlokpolitikOgNye2020 [315]
- most frequent ministry values

This does not only add a new unit of analysis to research on Northatlantic representation, but 

less about topics, more about hard numbers/numerical influence

- voting records: ungewöhnlich, aber aufgrund öffentlichkeitswirksamkeit angebracht. "Desuden er de parlamentariske spørgsmål en interessant variabel, fordi de giver et indblik i, hvilke problematikker medlemmerne selv har valgt at sætte på dagsordenen" [@harderFaeroskeOgGronlandske2022, 9] > Argumentation für/gegen roll call data?

  - Harder: looks at committees/questions, as it's custom to use committee membership of indicator of political activities  [@cleggParliamentaryRepresentationOverseas2022, 238] because that's where stuff happens [@hansenCommitteeAssignmentPolitics2010, 382] and most relevant for oversight [@westPoliticalPartiesParliamentary2025, 967] > why voting behavior instead
  - voting: not so influential [@arterNordicParliamentsMuchofamuchness2016, 214; @nannestadPolitischeSystemDanemarks2009, 82] and individual behavior not so relevant due to high party cohesion [@ejnarhansenReconsideringPartyDistances2008, 266; @hansenRealBorgenDanish2020, 3], but in the topic of voting behavior lies a lot of visibility/publicity

# Dataset Description

I am using the [*nordatlantisk-ft*](https://mtwente.github.io/nordatlantisk-ft) dataset which contains voting records of Folketinget members from Greenland and the Faroe Islands. Since Folketinget does not provide ready-to-use open datasets but metadata only [@berntzenParliamentaryOpenData2019, 9], I had to create both workflow and dataset myself. The dataset is the result of a pipeline in R, taking a list of MPs as input and then scraping voting records and metadata from Folketinget's Open Data Platform [@folketingetFolketingetsAbneData2026]. Both workflow and dataset are version-controlled using a git-based template including long-term archival on Zenodo [@mahrOneTemplateRule2025a], and are accompanied by extensive metadata and documentation to ensure reproducibility. [Codebook](https://mtwente.github.io/nordatlantisk-ft/docs/codebook.html) and [source code](https://github.com/mtwente/nordatlantisk-ft) are available via GitHub. 

For my analysis, I am going to use the data available as of `r format(file.info(here("data", "processed", "northatlantic_ft.rds"))$mtime, "%B %d, %Y")` [@twenteNordatlantiskft2026]. At this point, *nordatlantisk-ft* comprises `r length(unique(northatlantic_ft$vote_id))` records of votes cast by Northatlantic MPs between `r format(min(northatlantic_ft$date), "%B %d, %Y")` and `r format(max(northatlantic_ft$date), "%B %d, %Y")`[^valg26]. The dataset also contains metadata on each record -- e.g. date and topic of each vote or the overall roll call result, covering votes on `r length(unique(northatlantic_ft$roll_call_id))` proposals in Folketinget.

[^valg26]: PM Mette Frederiksen announced the date for the next general election to Folketinget (`r format(as.Date("2026-03-24"), "%B %d, %Y")`) on `r format(as.Date("2026-02-26"), "%B %d, %Y")`. The data used for this analysis thus does not contain records on voting behavior during the election campaign.

## MPs {.unnumbered .unlisted}

Since 1953, two MPs are elected in both Greenland and the Faroe Islands to join the 175 MPs from Denmark proper in Folketinget. Before the 1953 reform, Greenland did not have any parliamentary representation in Folketinget, as opposed to the Faroe Islands with one guaranteed seat since 1849 [@harderFaeroskeOgGronlandske2022, 8-9; @kjaerDanskeFolketingsmedlemmerParlamentarisk2004]. According to @harderFaeroskeOgGronlandske2022 [7], it is usually experienced politicians that are elected to represent the Faroe Islands constituency[^study], even though the job apparently has low priority compared to holding office 'at home' [@ackrenExploringRoleRegional2019, 22; cf. @harderCrossparliamentaryControl2025, 8; @johannesenUniHolmJohannesen2016].

[^study]: No comparable study exists for MPs from Greenland apart from the short MP profiles available in Danmarks Nationalleksikon [@billeFaeroskeFolketingsmedlemmer2025; @billeGronlandskeFolketingsmedlemmer2025].

In the time frame covered by Folketinget's Open Data Platform, there have been `r nrow(mp_names)` individual politicians from Greenland or the Faroe Islands who served as MPs (@tbl-MPs). `r sum(mp_names$origin == "GL")` of them are members of Greenlandic parties, and `r sum(mp_names$origin == "FO")` are representing Faroese parties. For MPs representing Greenland resp. the Faroe Islands in election periods earlier than `r format(min(northatlantic_ft$date), "%B %Y")`, see the list assembled by @harderSupplerendeMaterialeFaeroske2022.

`r sum(mp_names$substitute == TRUE)` of the politicians that are part of this dataset joined Folketinget in the role of a substitute member[^subst], filling in for one of the `r sum(mp_names$substitute == FALSE)` originally elected MPs while they were on leave [cf. @harderFaeroskeOgGronlandske2022, 12-13]. @fig-mandates shows membership periods and substitute members on a timeline for the duration that is covered by this dataset. Earlier membership periods of Edmund Joensen, Lisbeth Petersen, Høgni Hoydal and Tórbjørn Jacobsen are cut off from the timeline since there are no voting records available online earlier than `r format(min(northatlantic_ft$date), "%B %d, %Y")`.

[^subst]: Doris Jakobsen (SIU), Lisbeth Petersen (B), Magni Arge (E) and Sjúrður Skaale (C) all first became MPs as substitute members, but later joined as regular MPs after a general election.

```{r}
#| echo: false
#| message: false
#| label: fig-mandates
#| fig-height: 5
#| fig-cap: Timeline of MPs from Greenland and the Faroe Islands during the timeframe covered by this dataset. Substitutions shorter than three weeks are plotted as dots for better visibility.

# remove pre-2000 MPs (Edmund Joensen's and Lisbeth Petersen's first terms)
MP_dates <- MP_dates %>%
  filter(end >= as.Date("2000-01-01")) %>%
  mutate(origin = factor(origin, levels = c("GL", "FO")))

# Order MPs by first start date for nicer plotting
MP_dates <- MP_dates %>%
  group_by(origin, plotting_name) %>%
  mutate(first_start = min(start)) %>%
  ungroup()

# build facet-specific ordering
levels_gl <- MP_dates %>%
  filter(origin == "GL") %>%
  distinct(plotting_name, first_start) %>%
  arrange(desc(first_start)) %>%
  pull(plotting_name)

levels_fo <- MP_dates %>%
  filter(origin == "FO") %>%
  distinct(plotting_name, first_start) %>%
  arrange(desc(first_start)) %>%
  pull(plotting_name)

MP_dates <- MP_dates %>%
  mutate(
    plotting_name = case_when(
      origin == "GL" ~ factor(plotting_name, levels = levels_gl),
      origin == "FO" ~ factor(plotting_name, levels = levels_fo)
    )
  )

ggplot(MP_dates, aes(x = start,
                     xend = end,
                     y = plotting_name,
                     yend = plotting_name,
                     color = substitute)) +

  # for correct order of MPs across facets and scales
  geom_blank(aes(y = plotting_name)) +

  # Segments for all regular MPs and long-term substitutions
  geom_segment(
    data = MP_dates %>%
      filter(
        substitute == FALSE |
          (substitute == TRUE & !is.na(end) & as.numeric(end - start) >= 21)
      ),
    linewidth = 4,
    lineend = "butt"
  ) +

  # Points for short-term substitutions
geom_point(
  data = MP_dates %>%
    filter(
      substitute == TRUE &
        !is.na(end) &
        as.numeric(end - start) < 21
    ),
  aes(
    x = start,
    fill = substitute
  ),
  shape = 21,
  size = 1.5,
  stroke = 0.2,
  color = "#777777"
  ) +

  scale_color_manual(
    values = c("FALSE" = "#6195CF", "TRUE" = "#F7CB45"),
    labels = c("Regular", "Substitute")
  ) +
  scale_fill_manual(
  values = c("FALSE" = "#6195CF", "TRUE" = "#F7CB45"),
  guide = "none"
  ) +
  
  labs(color = "Membership") +

  # Folketing elections
  geom_vline(
    data = elections_ft,
    aes(xintercept = as.Date(date), linetype = election_type),
    color = "#000000",
    linewidth = 0.2
  ) +
  scale_linetype_manual(
    values = c("Folketing Election" = "longdash"),
    name = ""
  ) +

  facet_grid(
    origin ~ .,
    scales = "free_y",
    space = "free_y",
    labeller = labeller(origin = c("GL" = "Greenland", "FO" = "Faroe Islands"))
  ) +

  coord_cartesian(xlim = c(min(northatlantic_ft$date), NA)) +

  theme_nordatlantisk_basis() +
  theme(
    legend.position = "bottom",
    panel.grid.major.y = element_line(color = "#777777", linewidth = 0.05),
    panel.grid.minor.x = element_blank()
  )
```

## Political Parties {.unnumbered .unlisted}

Since 2004, MPs from the Faroe Islands were members of `r n2w(sum(political_parties$origin == "FO"))` different political parties. Voters in Greenland elected politicians from `r n2w(sum(political_parties$origin == "GL") - 2)` different political parties. However, two of the Greenlandic representatives -- Aleqa Hammond and Aki-Matilda Høegh-Dam -- switched party affiliation during their time as member of Folketinget and joined two relatively young parties (Nunatta Qitornai and Naleraq[^3]).

[^3]: Both parties have been formed quite recently -- Naleraq in 2014 and Nunatta Qitornai in 2017. See @ahlnessNunattaQitornaiParty2020 for general information on Nunatta Qitornai and their policies. Nunatta Qitornai has de facto been dissolved after the 2021 Inatsisartut election [@vestergaardhviidVittusQujaukitsoqMelder2025].

The Northatlantic MPs usually are members of the two biggest political parties of Greenland resp. the Faroe Islands [@elklitElectoralSystemFair2020]. These parties then act as de-facto ethnic regional parties in Folketinget [@ackrenExploringRoleRegional2019]. Before political parties were established in Greenland in the 1970s [@billeGronlandskeFolketingsmedlemmer2025; @michelsenGronlandPartisystemUdvikling1979], MPs were elected independently. Since then, one of the two Greenlandic seats has been held by social democratic *Siumut* (SIU) for most of the time. Only recently has Siumut not been represented in Folketinget, due to the MPs in question switching party affiliations (see @fig-party-timeline). The 'second' Greenlandic seat was affiliated with centre-right *Atassut* (A) from 1977 to 2001. Since then, these MPs have been members of socialist *Inuit Ataqatigiit* (IA). This dominance by two political parties is in contrast to the Faroese situation, where the four biggest parties take turns more often [@billeFaeroskeFolketingsmedlemmer2025].

All represented political parties are listed in the [Appendix](#list-of-parties) with additional information (where available) on their positions from the ParlGov dataset [@doringParlGov2024Release2024]. @fig-party-timeline shows the allocation of Folketinget seats to political parties from Greenland and the Faroe Islands on a timeline. At no point in time were both MPs of Greenland resp. the Faroe Islands members of the same party. In the time frame covered by this dataset, the Faroe Islands have been represented by one left-wing and one right-wing political party in Folketinget except for one election period (2015--2019), as opposed to the Greenlandic MPs having belonged to two left-wing parties.

The Northatlantic MPs have sometimes been members of Danish parliamentary groups. Over the years, this has varied between no membership, informal association to full memberships [@harderFaeroskeOgGronlandske2022, 17-18]. From 2001 until 2011, three Northatlantic MPs formed a separate parliamentary group ['Den Nordatlantiske Gruppe i Folketinget', @gadGronlandskeFolketingsmedlemmer2022; @harderFaeroskeOgGronlandske2022, S6].

```{r}
#| echo: false
#| message: false
#| label: fig-party-timeline
#| fig-cap: Timeline of political parties from Greenland and the Faroe Islands with an MP in Folketinget since 2004. Dashed lines indicate election dates.

# collapse MP_dates timeline into non-overlapping party intervals
party_periods <- MP_dates %>%
  select(party, origin, start, end) %>%
  arrange(party, origin, start, end) %>%
  group_by(party, origin)

# join colours from political_parties
party_periods <- party_periods %>%
  left_join(political_parties %>% select("party", "full_name", "colour"), by = "party")

# sort parties with start date
party_periods <- party_periods %>%
  group_by(origin) %>%
  mutate(party = factor(party, levels = rev(unique(party[order(start)])))) %>%
  ungroup()

# set party name labels including UF
y_labels <- setNames(party_periods$full_name, party_periods$party)
y_labels <- c(y_labels, "UF" = "No Party Affiliation")

# set UF colour to grey
party_periods$colour[party_periods$party == "UF"] <- "#777777"

ggplot(party_periods, aes(x = start, xend = end, y = party, yend = party, color = colour)) +
  geom_segment(linewidth = 8, lineend = "butt") +
  scale_color_identity() +

  scale_y_discrete(labels = y_labels) +
  
  # add Folketing Election dates
  geom_vline(data = elections_ft, aes(xintercept = as.Date(date), linetype = election_type),
             color = "#000000", linewidth = 0.2) +
  scale_linetype_manual(values = c("Folketing Election" = "longdash"), name = "") +
  
  facet_grid(origin ~ ., scales = "free_y", space = "free_y",
             labeller = labeller(origin = c("FO" = "Faroe Islands", "GL" = "Greenland"))) +
  
  coord_cartesian(xlim = c(min(northatlantic_ft$date), NA)) +
  
  theme_nordatlantisk_basis() +
  theme(
    legend.position = "none",
    panel.grid.major.y = element_line(color = "#777777", linewidth = 0.05),
    panel.grid.minor.x = element_blank()
  )
```

## Voting Records {.unnumbered .unlisted}

The observations available in the dataset contain voting decisions by the four MPs representing Greenland resp. the Faroe Islands at the time of each roll call vote. The final dataset used in this analysis comprises 22 variables that are explained in detail in the [codebook](https://mtwente.github.io/nordatlantisk-ft/docs/codebook.html). It contains three main categories of information for each observation: There is information on each recorded vote by an MP including the actual type of vote cast as well as metadata on the origin (`GL`/`FO`) and the party affiliation of the politician in question. Then there is metadata on the roll call vote such as the overall voting result and the topic of each proposal including the responsible ministry. And finally, the dataset also contains variables that provide information on the cabinets in Denmark and Greenland resp. the Faroe Islands at the time of the observation, e.g. whether the MP's party is part of the cabinet or not.

```{r bad-roll-calls}
#| echo: false
#| message: false

# Count how many times each ballot_id appears
counts <- table(northatlantic_ft$roll_call_id)

# Find roll_call_id values that do NOT appear exactly 4 times
counts_df <- as.data.frame(counts)
bad_roll_call_id <- as.data.frame(names(counts[counts != 4]))

# calculate share of total observations
error_share <- (nrow(bad_roll_call_id) + sum(!roll_call_results_ft$roll_call_id %in% northatlantic_ft$roll_call_id))/nrow(northatlantic_ft)
```

There are a few observations where data on Folketinget's Open Data Platform is unavailable or incomplete: For `r sum(!roll_call_results_ft$roll_call_id %in% northatlantic_ft$roll_call_id)` roll call votes, there are no individual voting records available at all. And for `r n2w(nrow(bad_roll_call_id))` roll call votes, there are only three registered observations for Northatlantic MPs instead of four[^missing]. In total, these errors correspond to `r scales::percent(error_share, accuracy = 0.01)` of all data.

[^missing]: This might be due to a missing functionality on Folketinget's Open Data Platform, where it is not possible to query (former) members based on their constituency. Maybe I missed a short-term temporary replacement MP. However, since I was not able to find any hint about MPs that are not yet part of the dataset in other sources, this maybe really just is a database error.

```{r}
#| echo: false
#| message: false
#| label: fig-n-rollcall
#| fig-cap: Number of roll call votes per year (excluding 2004 and 2026 due to incomplete coverage in the dataset).

##| fig-height: 1.2

rollcalls_per_year <- northatlantic_ft %>%
  mutate(year = lubridate::year(date)) %>%
  filter(year != 2026) %>% # filtering out 2026 because it will distort
  filter(year != 2004) %>% # filtering also 2004 because records only start in October
  group_by(year) %>%
  summarise(
    n_rollcalls = n_distinct(roll_call_id),
    .groups = "drop"
  )

# compute global maximum for y-axis
y_max <- max(rollcalls_per_year$n_rollcalls, na.rm = TRUE) + 1 


ggplot(
  rollcalls_per_year,
  aes(x = factor(year), y = n_rollcalls)
) +
  
  geom_col(width = 0.7, fill = "#6195CF") +
  
  scale_y_continuous(
    limits = c(0, y_max),
    breaks = c(seq(0, 600, 200), 700),
    labels = c(seq(0, 600, 200), 700),
    expand = expansion(mult = c(0, 0.01))
  ) +
  
  theme_nordatlantisk_basis() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.ticks.x = element_blank(),
    axis.line.y.left = element_blank(),
    panel.grid.major.y = element_line(color = "#777777", linewidth = 0.05)
  )
```

```{r voting-blocks}
#| echo: false
#| message: false
#| warning: false
# calculate share of roll call votes per block
blocks <- as.data.frame(table(northatlantic_ft$gvt_bloc_dk))
```

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-vote-types
#| tbl-cap: "Types of roll call votes."
#| tbl-cap-location: bottom

roll_call_summary <- northatlantic_ft %>%
  filter(roll_call_type_id %in% c(1, 2, 3, 4)) %>%
  count(roll_call_type_id, name = "n") %>%
  mutate(
    Type = case_when(
      roll_call_type_id == 1 ~ "Final Adoption",
      roll_call_type_id == 2 ~ "Committee Recommendation",
      roll_call_type_id == 3 ~ "Proposed Adoption",
      roll_call_type_id == 4 ~ "Proposed Amendment"
    )
  ) %>%
  select(ID = roll_call_type_id, Type, n)

kable(
  roll_call_summary,
  col.names = c("ID", "Type", "Count"),
  booktabs = TRUE,
  format = "latex"
)
```

During the timeframe covered by this dataset, there are `r floor(mean(rollcalls_per_year$n_rollcalls))` roll call votes in Folketinget per year on average, with a maximum number of `r max(rollcalls_per_year$n_rollcalls)` votes in `r rollcalls_per_year %>% slice_max(n_rollcalls, n = 1) %>% pull(year)` (@fig-n-rollcall). The dataset covers all types of roll call votes in Folketinget (@tbl-vote-types): `r scales::percent(length(northatlantic_ft$roll_call_type_id[northatlantic_ft$roll_call_type_id == "1"])/nrow(northatlantic_ft), accuracy = 0.1)` of the votes are final decisions on a proposal, and `r scales::percent(length(northatlantic_ft$roll_call_type_id[northatlantic_ft$roll_call_type_id == "4"])/nrow(northatlantic_ft), accuracy = 0.1)` are votes on amendments to a proposal. Overall, `r scales::percent(mean(northatlantic_ft$roll_call_pass, na.rm = TRUE), accuracy = 0.1)` of the proposals pass with a positive voting result.

The dataset spans across across six full cycles of Folketinget elections (and the last few months of the 2001--2005 election cycle). Within this timeframe, there have been seven elections to Inatsisartut in Greenland and five elections to the Faroese Løgting. Of the roll call votes covered in the dataset, `r scales::percent(blocks$Freq[blocks$Var1 == "left"]/sum(blocks$Freq), accuracy = 0.1)` took place during a cabinet formed by the red/left block was in office in Denmark (during blue/right block coalitions: `r scales::percent(blocks$Freq[blocks$Var1 == "right"]/sum(blocks$Freq), accuracy = 0.1)`, and for the current cross-block coalition: `r scales::percent(blocks$Freq[blocks$Var1 == "across"]/sum(blocks$Freq), accuracy = 0.1)`). The ministry responsible for each proposal being voted on is also a variable in the dataset. The most records refer to the Ministry of Finance, the Ministry of Justice, and the Ministry of Employment (@tbl-top-ministries). This has to be read with a grain of salt, since there are a lot of `NA` values and because designations[^full] change frequently so that e.g. 'Klima- og Energiministeriet' and 'Klima-, Energie- og Bygningsministeriet' etc. are counted as different entities. 

[^full]: Refer to the codebook for a [full list](https://mtwente.github.io/nordatlantisk-ft/docs/codebook.html#northatlantic_ftministry) of all `r n_distinct(northatlantic_ft$ministry, na.rm = TRUE)` designations that appear in the dataset.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-top-ministries
#| tbl-cap: "Most frequent values for `ministry`."
#| tbl-cap-location: bottom

northatlantic_ft %>%
  count(ministry, sort = TRUE) %>%
  slice_head(n = 5) %>%
  knitr::kable(
    col.names = c("Ministry", "Count"),
    align = c("l", "r"),
    booktabs = TRUE,
    format = "latex"
  )
```

MPs can either vote for, against or neither for nor against a proposal. Additionally, absences are recorded. @fig-ft-vote-distribution shows the distribution of vote types per roll call vote for all MPs in Folketinget. At first glance, it stands out how the number of absent MPs hovers around 70 for all years covered by the timeline (cf. [Descriptive Statistics](#descriptive-statistics) in the Appendix). This is due to Folketinget being a 'working parliament', where MPs are regularly cleared from taking part in the plenary sessions which are "mostly poorly attended" [@arterNordicParliamentsMuchofamuchness2016, 214].

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-ft-vote-distribution
#| fig-env: "figure*"
#| fig-cap: Density plot visualising the number of votes cast by all MPs for each roll call vote by type.

make_density_plot <- function(data, y_var, title, low_col, high_col) {
  
  ggplot(data, aes(x = date, y = {{ y_var }})) +
    
    geom_bin2d(aes(fill = after_stat(count)), bins = 80) +
    scale_fill_gradient(low = low_col, high = high_col) +
    
    scale_x_date(
      breaks = as.Date(c("2005-01-01", "2015-01-01", "2025-01-01")),
      date_labels = "%Y",
      expand = expansion(mult = c(0.01, .05))
    ) +
    
    scale_y_continuous(
      limits = c(0, 179),
      labels = c(0, 50, 100, 150, 179),
      breaks = c(0, 50, 100, 150, 179)
    ) +
    
    ggtitle(title) +
    
    theme_nordatlantisk_basis() +
    theme(
      axis.title.y = element_blank(),
      axis.line.y.left = element_blank(),
      legend.position = "none",
      panel.grid.major.y = element_line(color = "#777777", linewidth = 0.05),
      plot.title = element_text(size = 8),
      axis.text.x = element_text(size = rel(0.7))
    )
}

scatterplot_for <- make_density_plot(northatlantic_ft, ft_for, "for", "#6195CF", "#436890")

scatterplot_against <- make_density_plot(northatlantic_ft, ft_against, "against", "#F7CB45", "#ac8e30")

scatterplot_abstention <- make_density_plot(northatlantic_ft, ft_abstention, "abstention", "#777777", "#535353")

scatterplot_absent <- make_density_plot(northatlantic_ft, ft_absent, "absent", "#D1BBD7", "#928296")

patchworked_plot <- scatterplot_for + scatterplot_against + scatterplot_abstention + scatterplot_absent + plot_layout(ncol = 4, axes = "collect")

patchworked_plot
```

Looking at the North Atlantic MPs' voting records, it is striking how few times they actually cast a vote: Filtering for those cases results in `r length(northatlantic_ft$vote_type_id[northatlantic_ft$vote_type_id != "3"])` non-absent observations (out of `r nrow(northatlantic_ft)` cases), translating to `r mean(northatlantic_ft$vote_type_id == "3") %>% scales::percent(accuracy = 0.1)` of observations being coded as 'absent' (@fig-absence-share) -- an observation that corresponds well to the findings of previous research. Of the non-absent records, `r scales::percent(sum(northatlantic_ft$vote_type_id == 1, na.rm = TRUE) / sum(northatlantic_ft$vote_type_id != 3, na.rm = TRUE), accuracy = 0.1)` are votes in favor, `r scales::percent(sum(northatlantic_ft$vote_type_id == 2, na.rm = TRUE) / sum(northatlantic_ft$vote_type_id != 3, na.rm = TRUE), accuracy = 0.01)` are votes against, and `r scales::percent(sum(northatlantic_ft$vote_type_id == 4, na.rm = TRUE) / sum(northatlantic_ft$vote_type_id != 3, na.rm = TRUE), accuracy = 0.01)` are votes neither for nor against.

```{r, fig.pos = 'H'}
#| echo: false
#| message: false
#| fig-height: 1.2
#| label: fig-absence-share
#| fig-cap: Northatlantic MPs are listed as absent in 98% of the observations in this dataset.

# #| fig-env: "figure*" removed temporarily

# Compute proportions of vote types
vote_props <- northatlantic_ft %>%
  count(vote_type_id) %>%
  mutate(
    share = n / sum(n),
    vote_type = factor(vote_type_id,
                       levels = c(3, 4, 2, 1),
                       labels = c("absent", "abstention", "against", "for"))
  )

# Single horizontal bar plot
ggplot(vote_props, aes(x = 1, y = share, fill = vote_type)) +
  geom_col(width = 0.5) +
  coord_flip() +
  scale_fill_manual(
    name = "Vote",
    labels = c("for", "against", "abstention", "absent"),
    values = c("#6195CF", "#F7CB45", "#777777", "#D1BBD7"),
    breaks = c("for", "against", "abstention", "absent")
  ) +

  scale_y_continuous(
    labels = scales::label_percent(),
    breaks = c(0, 0.05, 0.25, 0.50, 0.75, 1.00),
    expand = expansion(mult = c(0, .015))
  ) +
  
  labs(x = NULL, y = NULL) +
  theme_nordatlantisk_basis() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.ticks.x = element_line(color = "#000000", linewidth = 0.1),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.position = "top"
  )
```

The level of participation in roll call votes (i.e. casting a vote either for, against, or neither for nor against) varies over time. Overall, rather low proportions of non-absent votes -- rarely exceeding 3% -- are most common in the dataset (@fig-vote-share-per-year). However, the share of votes cast among MPs from the Faroe Islands increased in recent years (2023 and 2025, but not 2024). This observation is not true for the Greenlandic MPs, where a single peak is recorded in the data for 2015. The latter can be attributed to (then) newly elected MP Aaja Chemnitz (Inuit Ataqatigiit) while the majority of the votes recorded from Faroese MPs since 2023, a remarkable increase in share of non-absent records, were cast by MP Anna Falkenberg (Sambandsflokkurin) who was elected to Folketinget at the end of 2022.

It is self-evident that the MPs with the most observations in the dataset are those who are/were the longest-sitting members (@fig-votes-per-mp) -- Sjúrður Skaale (FO/C, `r sum(northatlantic_ft$MP_id == 262, na.rm = TRUE)` records), Edmund Joensen (FO/B, `r sum(northatlantic_ft$MP_id == 247, na.rm = TRUE)` records) and Aaja Chemnitz (GL/IA, `r sum(northatlantic_ft$MP_id == 15757, na.rm = TRUE)` records). As MPs are absent for most roll calls, @fig-votes-per-mp-filtered shows the distribution only with votes that were actually cast. Long-term membership, however, does not necessarily correspond with a larger number in registered votes, as demonstrated by Sjúrður Skaale's share of `r scales::percent(sum(northatlantic_ft$vote_type_id[northatlantic_ft$MP_id == 262] != 3) / length(northatlantic_ft$vote_type_id[northatlantic_ft$MP_id == 262]), accuracy = 0.01)` non-absent records.

::: {#fig-vote-share-per-year}
<!-- in a fenced div because of the dynamically updating caption -->

```{r}
#| echo: false
#| message: false

votes_per_year_origin <- northatlantic_ft %>%
  mutate(year = lubridate::year(date)) %>%
  group_by(origin, year) %>%
  summarise(
    share_present = mean(vote_type_id != 3),
    .groups = "drop"
  )

# compute global maximum for y-axis
y_max <- max(votes_per_year_origin$share_present, na.rm = TRUE) + 0.001 

# faceted bar plots
ggplot(
  votes_per_year_origin,
  aes(x = factor(year), y = share_present)) +

  geom_col(width = 0.7, fill = "#6195CF") +
  
  scale_y_continuous(
    limits = c(0, y_max),
    labels = scales::label_percent(accuracy = 1),
    expand = expansion(mult = c(0, 0.01))
  ) +
  
  facet_grid(
    origin ~ .,
    scales = "fixed",
    space = "fixed",
    labeller = labeller(origin = c("GL" = "Greenland", "FO" = "Faroe Islands"))
  ) +
  
  theme_nordatlantisk_basis() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.ticks.x = element_blank(),
    axis.line.y.left = element_blank(),
    panel.spacing.y = unit(1, "lines"),
    panel.grid.major.y = element_line(color = "#777777", linewidth = 0.05)
  )
```

Share of non-absent records per year (for, against, abstaining) as of `r format(file.info(here("data", "processed", "northatlantic_ft.rds"))$mtime, "%B %Y")`.
:::

```{r}
#| echo: false
#| message: false
#| label: fig-votes-per-mp
#| fig-cap: Full records coded by the type of vote cast (for, against, abstaining, absent).

plot_votes <- ggplot(northatlantic_ft_ordered, aes(y = plotting_name, fill = vote_type_id)) +
  geom_bar() +
  scale_x_continuous(expand = expansion(mult = c(0, .05))) +
  scale_y_discrete() +
  scale_fill_manual(
    name = "Vote",
    labels = c("for", "against", "abstention", "absent"),
    values = c("#6195CF", "#F7CB45", "#777777", "#D1BBD7"),
    breaks = c("1", "2", "4", "3")
  ) +

  facet_grid(origin ~ ., scales = "free_y", space = "free_y",
            labeller = labeller(origin = c("GL" = "Greenland", "FO" = "Faroe Islands"))
            ) +
    
  theme_nordatlantisk_basis()

plot_votes
```

```{r}
#| echo: false
#| message: false
#| label: fig-votes-per-mp-filtered
#| fig-cap: Number of votes cast coded by the type of vote (for, against, abstaining). Includes only data with the MP in attendance.

plot_votes_without_absences <- ggplot(northatlantic_ft_without_absences, aes(y = plotting_name, fill = vote_type_id)) +
  geom_bar() +
  scale_x_continuous(expand = expansion(mult = c(0, .05))) +
  scale_y_discrete() +
  scale_fill_manual(
    name = "Vote",
    labels = c("for", "against", "abstention"),
    values = c("#6195CF", "#F7CB45", "#777777")
  ) +

  facet_grid(origin ~ ., scales = "free_y", space = "free_y",
          labeller = labeller(origin = c("GL" = "Greenland", "FO" = "Faroe Islands"))
          ) +

  theme_nordatlantisk_basis()

plot_votes_without_absences
```

# Results

<!-- - per hypothesis -->

## Implications, Future Research

## Critical Assessment

<!-- * intro&conclusion should speak to each other/mirror each other -->

use of roll call votes as method

<!-- conclusion paragraph structure: refer to powner handbook again -->

# References {.unnumbered}

::: {#refs}
:::

\onecolumn

# Appendix {.appendix}

## List of MPs {.appendix .unnumbered .unlisted}

Since Folketinget's Open Data Platform does not provide a function to search for (former) MPs per constituency, this list was compiled based on previous work [e.g. @harderSupplerendeMaterialeFaeroske2022; @gadGronlandskeFolketingsmedlemmer2022; @billeFaeroskeFolketingsmedlemmer2025; @billeGronlandskeFolketingsmedlemmer2025] and newspaper reports, as well as by searching for strings like 'grønland\*' and 'færøsk\*' on the Open Data Platform. This is not a very efficient way of searching since it will also return all MPs who at one point were members of Folketinget's commissions on Greenland or the Faroe Islands, or, additionally, all MPs who mention one of these places in their (free text) biography (this concerns e.g. former workplaces etc.). If there should be any MP still missing from this list, it is however easy to add them to the dataset and to reproduce the workflow.

The nordatlantisk-ft repository also contains complementary metadata on the MPs in this dataset, amongst other things a [complete membership timeline](https://github.com/mtwente/nordatlantisk-ft/blob/21774728bf1cd50901ba2b2fe0341018cc9c6ac9/data/processed/csv/MP_dates.csv) including temporary replacements, and a data frame [linking](https://github.com/mtwente/nordatlantisk-ft/blob/21774728bf1cd50901ba2b2fe0341018cc9c6ac9/data/processed/csv/MP_names.csv) each MP to the corresponding Wikidata item from where further information can be inferred.

```{r}
#| echo: false
#| message: false
#| label: tbl-MPs
#| tbl-colwidths: [12, 30, 25, 10, 10, 13]

MPs <- mp_names %>%
  select(-wikidata_id) %>%
  select(origin, everything()) %>%
  arrange(origin, surname)

printable_names <- c("Origin", "Surname(s)", "First Name(s)", "MP ID", "Party", "Substitute")

kable(MPs, caption = "List of MPs from Greenland and the Faroe Islands in Folketinget (October 2004 -- January 2026).", col.names = printable_names)
```

## List of Parties {.appendix .unnumbered .unlisted}

Information on each party's positions are taken from the 2024 ParlGov release [@doringParlGov2024Release2024]. ParlGov does, however, not include metadata on the two Greenlandic parties that have been represented in Folketinget only recently -- Nunatta Qitornai in 2018/2019 and, since 2025, Naleraq. Additionally, Aleqa Hammond was MP without party affiliation from `r format(MP_dates$start[MP_dates$period_id == "hammond_2"], "%B %d, %Y")` to `r format(MP_dates$end[MP_dates$period_id == "hammond_2"], "%B %d, %Y")` [@twenteNordatlantiskft2026].

```{r}
#| echo: false
#| results: asis

parties_gl <- political_parties |>
  filter(origin == "GL") |>
  mutate(full_name_abbr = paste0(full_name, " (", party, "/", origin, ")")) |>
  select(
    full_name_abbr, party_family,
    left_right, state_market,
    liberty_authority, anti_pro_eu
  )

parties_fo <- political_parties |>
  filter(origin == "FO") |>
  mutate(full_name_abbr = paste0(full_name, " (", party, "/", origin, ")")) |>
  select(
    full_name_abbr, party_family,
    left_right, state_market,
    liberty_authority, anti_pro_eu
  )

# dirty fix for visualisation purposes
parties_gl  <- parties_gl  |> mutate(across(everything(), as.character))
parties_fo  <- parties_fo  |> mutate(across(everything(), as.character))

combined <- bind_rows(
  parties_gl,
  parties_fo
)

colnames(combined) <- c(
  "Party", "Party Family",
  "Left--Right", "State--Market",
  "Liberty--Authority", "Anti--Pro-EU"
)

#knitr::kable(
#    combined,
#    align = "llcccc",,
#    booktabs = TRUE,
#    format = "latex"
#  )

stargazer(
  combined,
  summary = FALSE,
  rownames = FALSE,
  header = FALSE,
  column.sep.width = "2pt",
  type = "latex",
  float = FALSE,
  title = "Political parties from Greenland and the Faroe Islands with an MP in the Folketing since 2004.",
  label = "tbl-parties"
)
```

## Descriptive Statistics {.appendix .unnumbered .unlisted}

For `r n2w(sum(is.na(northatlantic_ft$ft_for)))` roll call votes, no individual voting results are available on Folketinget's Open Data Platform, explaining the mismatch between this table's `N` value and the total number of observations in the dataset (`r nrow(northatlantic_ft)`). Refer to the [online documentation](https://mtwente.github.io/nordatlantisk-ft/docs/codebook.html) with a comprehensive codebook to retrace the workflow steps that are needed to calculate this data structure.

```{r}
#| echo: false
#| output: asis

northatlantic_numeric <- northatlantic_ft |>
  select(ft_for, ft_against, ft_abstention, ft_absent)

stargazer(as.data.frame(northatlantic_numeric), header = FALSE, digits = 2, type = "latex", digit.separator = "'", summary.stat = c("n", "min", "max", "mean", "median", "p25", "p75", "sd"), label = "tbl-desc-data", title = "Descriptive statistics for the numerical variables in the dataset.", float = FALSE)
```

## Word Count {.appendix .unnumbered .unlisted}

Currently featuring {{wordcount}} words without references. Including references, there are {{wordcountref}} words in this document.

\twocolumn

<!-- ursprüngliche notizen aus dem Notizbuch:

H1 MPs stimmen eher ab, wenn Mehrheitsverhältnisse knapp sind (nachträglich¿)
H2 MPs stimmen eher ab, wenn ihre Partei in FO/GL nicht Teil des Kabinetts ist
H3 Ob MPs abstimmen, hängt nicht davon ab, ob ihre Partei demselben Block angehört wie die aktuelle Regierung in DK

Themen-bezogene Hypothesen: Schwierig, das sinnvoll zu modellieren in angemessenem Aufwand


Attendance != active participation? Difference in vote type 3 and 4


H4 nordatlantiske mandated can be understood as a „single block“/exhibit a cohesive voting pattern (controlling for bloc membership?) see Hix 2006 index, Hug 2013 p. 147
-->