---
title: "Counting to 90 in Greenlandic and Faroese"
subtitle: Analysing the North Atlantic MPs Voting Behaviour in the Danish Parliament
authors:
  - name: Moritz Twente
    affiliation: University of Basel
    email: moritz.twente@unibas.ch
    orcid: 0009-0005-7187-9774
abstract: > 
    nordatlantisk-ft is compiled from data available on Folketingets Open Data Platform [@folketingetFolketingetsAbneData2026] and comprises 40899 records of votes cast by Northatlantic MPs as per the most recent update (February 06, 2026). The dataset documents voting behavior of MPs from Greenland and the Faroe Islands and also contains metadata on each record – e.g. date and topic of each vote or the overall voting result, covering votes on 10227 proposals in Folketinget. Refer to the [documentation](https://mtwente.github.io/nordatlantisk-ft) for details.
date: last-modified
date-format: long
bibliography: ../assets/nordatlantisk-ft.bib
#csl: ../assets/american-political-science-association.csl
reference-section-title: References
link-citations: true
lang: en
output-file: paper.pdf
filters:
  - wordcount
wordcount:
  count-code-blocks: false
  count-inline-code: false
format: 
  pdf: 
    documentclass: article
    classoption: twocolumn
    mainfont: Jost
    filters: [../assets/quotation-marks.lua]
    quot-lang: de-CH
    colorlinks: true
    toc: true
    toc-depth: 2
    geometry:
      - top=22mm
      - bottom=18mm
    include-in-header:
      text: |
        \usepackage{fancyhdr}
        \pagestyle{fancy}
        \fancyhf{} % clear all header and footer fields

        % first page footer
        \fancypagestyle{plain}{
        \fancyhf{}
        \fancyfoot[L]{\footnotesize Moritz Twente, 19-911-940, moritz.twente@unibas.ch}
        \fancyfoot[R]{\footnotesize \thepage}
        \renewcommand{\headrulewidth}{0pt}
        }

        % Right-justified header (smaller font)
        \fancyfoot[L]{\footnotesize Moritz Twente, 19-911-940, moritz.twente@unibas.ch}

        % Page number in footer (centered)
        \fancyfoot[R]{\footnotesize \thepage}

        % Remove header line (optional, cleaner look)
        \renewcommand{\headrulewidth}{0pt}

        % reduce whitespace below header
        \setlength{\headsep}{9pt}
---

<!-- 8000 -- 10'000 words, 20--25 pages excl. Bibliography -->

<!-- writing: 
- every paragraph includes one argument/thought
- structure of a paragraph: topic sentence, development of the argument (2-3 sentences that elaborate your idea, e.g. by example)
- concluding sentence, link to the next paragraph -->

```{r setup}
#| echo: false
#| message: false

# Setup -----
## Packages -----

library(here)
library(ggplot2)
library(dplyr)
library(readr)
library(knitr)
library(forcats)
library(lubridate)
library(scales)
library(showtext)
library(xfun)

## Install Font -----

font_add("Jost",
  regular = here("assets", "fonts", "ttf", "Jost-400-Book.ttf"))

showtext_auto()

## Read Data -----

### List of MPs -----
mp_names <- readRDS(here("data", "processed", "MP_names.rds"))

### List of political parties -----
political_parties <- readRDS(here("data", "processed", "political_parties.rds"))

### Voting Records Database -----
northatlantic_ft <- readRDS(here("data", "processed", "northatlantic_ft.rds"))

### Timeline of Mandates -----
MP_dates <- readRDS(here("data", "processed", "MP_dates.rds"))
elections <- readRDS(here("data", "processed", "election_dates.rds"))

#### Convert dates
MP_dates <- MP_dates %>%
  mutate(
    start = ymd(start),
    end   = ymd(end),
    # Replace NA in "end" with today's date
    end   = if_else(is.na(end), Sys.Date(), end)
  )

## Manipulate Data -----

### Avoid confusing MPs with the same last name
MP_dates <- MP_dates %>%
  mutate(
    plotting_name = case_when(
      MP_id == "12283" ~ "A. V. Johannesen",
      MP_id == "6687"  ~ "K. L. Johannesen",
      TRUE             ~ surname
    )
  )

northatlantic_ft <- northatlantic_ft %>%
  mutate(
    plotting_name = case_when(
      MP_id == "12283" ~ "A. V. Johannesen",
      MP_id == "6687"  ~ "K. L. Johannesen",
      TRUE             ~ surname
    )
  )

### Order origin factor levels for facet order
northatlantic_ft <- northatlantic_ft %>%
  mutate(origin = forcats::fct_rev(origin))

### Order dataset by MP surnames for plotting
northatlantic_ft_ordered <- northatlantic_ft %>%
  mutate(surname = fct_reorder(surname, vote_type_id, .fun = length)) %>%
  mutate(plotting_name = fct_reorder(plotting_name, vote_type_id, .fun = length)) %>%
  mutate(vote_type_id = fct_relevel(vote_type_id, "1", "2", "4", "3"))

### Subset only with votes that were actually cast
northatlantic_ft_without_absences <- subset(northatlantic_ft_ordered, vote_type_id != 3) %>%
  mutate(surname = fct_reorder(surname, vote_type_id, .fun = length)) %>%
  mutate(plotting_name = fct_reorder(plotting_name, vote_type_id, .fun = length))

### Subset Election Dates for Folketinget only

elections_ft <- elections %>%
  filter(election_type == "Folketing Election")

## Set Plotting Theme -----

theme_nordatlantisk_basis <- function(){
  
  theme_grey(base_family = "Jost") %+replace%
    theme(axis.title = element_blank(),
          axis.ticks.y = element_blank(),
          panel.background = element_blank(),
          panel.grid.major.x = element_line(color = "#000000", linewidth = 0.1),
          panel.grid.minor.x = element_line(color = "#000000", linewidth = 0.1),
          panel.grid.major.y = element_blank(),
          panel.grid.minor.y = element_blank(),
          axis.ticks.x = element_line(color = "#000000", linewidth = 0.1),
          axis.line.y.left = element_line(color = "#000000", linewidth = 0.1))
}

```

# Introduction

<!-- Motivation / general topic / relevance -->

- history of memberships @kjaerDanskeFolketingsmedlemmerParlamentarisk2004

parties state they do not want to interfere in 'domestic' Danish politics [@kurrild-klitgaardElectionInversionsCoalitions2013, 130]

they are usually not official members of cabinet, except for Mikael Gam [@skjaevelandMindretalsregeringerDanmarkFor2003]

### DK politics

### Rigsfællesskabet

### Postcolonialism

### Geopolitics on the World Stage


- inner-Rigsfællesskabet debates very current,
  - mechanisms of Rigsfællesskabet dynamics, but within Folketinget?
- opens up international postcolonialism debates
  - "coming down from the ice" [@sjolieDanmarkManglerOpgor2008]
  -  "Greenlandic representatives have been bullied by representatives from Danish parties" [@harderArcticIndigenousRepresentation2024, 141] cf. @cleggParliamentaryRepresentationOverseas2022
- link to general geopolitics focus on Greenland
  + DK using GL/FO (and their FT membership) to legitimize themselves as Arctic state [@westDanishRealmArctic2025, 589]

* Übergreifend hat die Beachtung dieser Abgeordneten zugenommen, seit Grönland aus geopolitischer Sicht in den Fokus gerückt ist, und nicht zuletzt auch Dänemarks Status als Akteur in der Arktis legitimiert werden soll.
Ethnic minority representation (betrifft dann allerdings nur Grönland und nicht die Färöer)
* Dazu kommt noch die ganze post-colonialist-Perspektive ("vi har købt os en grønlænder" etc.).


 - research stating that they do not want to and are also expected to not to interfere

but is there more?

## Research Questions, Relevant X/Y variables

1. In general, when do the Northatlantic MPs take part in roll call votes?
2. The 2x2 MPs are conceived as one group by DK politicians and media. Do they also vote like one group?
3. When Northatlantic MPs have the chance to have influence, they act on it (based on decades of more or less anecdotal evidence). So, does the seat share of Danish coalition parties influence voting behaviour of Northatlantic MPs?
4. can the Northatlantic MP's voting behavior also be considered a form of cross-parliamentary control?

# Literature Review

<!-- - summary of relevant literature for your research question!
- aim of literature review: identify research gap (and by this your own contribution)
- do not mention every paper written about your case but the most important+related ones
- pathbreaking papers and latest developments should be included
- Direct citations only if absolutely needed, better are paraphrasing -->

Over the past decades, the activity of Folketing MPs from Greenland and the Faroe Islands has been written about time and again -- usually when there are narrow majorities and the 'Northatlantic' votes might become relevant to the voting outcomes, e.g. after elections in 2022, 2015, 2011 or 2007 [@harderFaeroskeOgGronlandske2022; @sorensenEfterBrokBla2022]. While situations like these are regularly reported on by journalists [e.g. @nathanGronlandskMferSer2024; @veirumPartilederePatalerGronlandsk2022; @frostLokkeRaserNordatlantiske2022; @skouKaareFireMandater2008][^zotero], the attention they receive from researchers has long been marginal.

[^zotero]: For a larger collection of reports in mostly Danish-language media from recent years, see my [nordatlantisk-ft Zotero collection](https://www.zotero.org/groups/5346749/nordatlantisk-ft).

If Greenlandic or Faroese political representation in Folketinget does appear in research output, it is often referred to in anecdotes [e.g. "A coalition saved by Greenland", @millerFriendsRivals1996, 80] or in parentheses [e.g. @klintDanishLegislatorsDatabase2023; @elklitElectoralSystemFair2020; @nannestadPolitischeSystemDanemarks2009; @kjaerDanskeFolketingsmedlemmerParlamentarisk2004], acknowledging its existence, but mostly not considering it for analysis implicitly or explicitly [e.g. @ejnarhansenReconsideringPartyDistances2008; @skjaevelandPartyCohesionDanish2001; @valenGeographyPoliticalRepresentation2000; @westPoliticalPartiesParliamentary2025; @elklitDenmarkSimplicityEmbedded2005]. Given the anecdotal evidence about the Northatlantic MPs' presence and the often heated reactions to situations where roll call results are depending on votes from (at least one of) them, it is striking how research related to Folketinget and its dynamics has largely not taken up on this aspect.

In the following paragraphs, I will take a closer look at literature on the functioning of Folketing itself, on coalition and government formation, on the Danish election system as well as on the political parties.

Writing about parliamentary cross-bloc cooperation, @green-pedersenBlokpolitikOgNye2020 [311] do not include political parties from Greenland and the Faroe Islands when counting the number of parties that are represented in parliament at all. Neither does @helboepedersenParliamentFOLKETINGETPowerful2020 in her investigation institutional powers of Folketinget, although she briefly mentions the existence of Northatlantic MPs. @hansenCommitteeAssignmentPolitics2010 states he is not considering these MPs in an analysis of committee assignments. In a categorisation of roll call votes, @skjaevelandPartyCohesionDanish2001 explicitly excludes matters concerning Northatlantic MPs from a category of domestic issues.

The results from a review of literature on coalition and government formation in Denmark are similar: @green-pedersenBlocPoliticsVs2005 and @skjaevelandModellingGovernmentFormation2009 find that gaining government power depends on the median legislator, often centre parties -- which in Denmark has traditionally been associated with Det Radikale Venstre [@green-pedersenDetRadikaleVenstres2003]. Northatlantic MPs are not touched upon in either paper, nor are they mentioned by @christiansenMinorityCoalitionGovernance2014 in their analysis of coalition agreements. @skjaevelandModellingGovernmentFormation2009, however, includes Northatlantic MPs in his research on median parties in Denmark, acknowledging their importance in some coalition life cycles. The take-away from his short digression on the North Atlantic Folketing is that "the more important their votes, the more involved they have become" [-@skjaevelandModellingGovernmentFormation2009, 723]. But even when describing situations where the voting behavior of North Atlantic MPs proved to be decisive for government formations, for some researchers they are still only partially relevant to the distribution of power in Danish politics [@skjaevelandMindretalsregeringerDanmarkFor2003].

Skjæveland 2003 ausführlicher beschreiben, vlg. Würdigung bei @harderFaeroskeOgGronlandske2022

*"I den politologiske litteratur er særligt de nordatlantiske mandaters intervention i forbindelse med regeringsdannelser udførligt beskrevet (Skjæveland, 2003). Ifølge Skjæveland, der undersøger perioden 1953-1998, har den generelle opfattelse af disse mandater været, at de ikke blander sig i (syd)dansk politik, omend der samtidig hersker en opfattelse af, at de i stigende grad er blevet involveret heri (Skjæveland, 2003: 114-115; Michelsen, 1979: 55)"*

With regard to the composition of Folketinget, @kurrild-klitgaardElectionInversionsCoalitions2013 analyses election inversions, meaning the party/political bloc winning the most votes does not end up with the most parliamentary seats after the election -- a situation the Northatlantic MPs have ended up in repeatedly. Using the 1971 election, he demonstrates that the lower cost of winning a Folketinget mandate in the Faroe Islands resulted in a margin so narrow that, if added to continental Danish overall results, would have resulted in an alternative majority [-@kurrild-klitgaardElectionInversionsCoalitions2013, 131]. This is, however, not considered to be a flaw, and the election system of Denmark is regarded to be exceptionally well-functioning [@elklitElectoralSystemFair2020; @lijphartForeword2005], even though that assessment, again, excludes elections in Greenland and the Faroe Islands [cf. @elklitDenmarkSimplicityEmbedded2005]. An otherwise comprehensive database with biographic information on Folketinget legislators by @klintDanishLegislatorsDatabase2023 also does not include anyone from Greenland and the Faroe Islands.

Research on political parties in Denmark, e.g. by @green-pedersenPartySystemOpen2020 or @pedersenPartyDistancesDanish1971, does not cover parties from Greenland or the Faroe Islands either. This is hardly surprising, given that the political parties competing in elections in Denmark, Greenland and the Faroe Islands are separate entities and each party can only be voted for from within either of the three parts of the Danish Realm [@kurrild-klitgaardElectionInversionsCoalitions2013]. Likewise, research on political parties of Greenland and the Faroe Islands features a domestic focus as well [e.g. @ahlnessNunattaQitornaiParty2020; @westSkipanarligarFortreytirOg2022; @michelsenGronlandPartisystemUdvikling1979]. Institutional associations between Greenlandic and Danish parties linked to parliamentary work in Folketinget are mentioned by @ackrenPoliticalPartiesGreenland2015, building on ideologically similar backgrounds of Siumut of Greenland and Socialdemokratiet in Denmark resp. of Atassut (GL) and Venstre (DK). However, Folketinget representation does not seem to be a central issue for regional parties in Greenland and the Faroe Islands. Because Greenland and the Faroe Islands only hold two seats each, @ackrenExploringRoleRegional2019 deem it unlikely that any Northatlantic proposal would gain enough traction in Folketinget anyway. For these regional parties, the preferred mode of conducting relations with Denmark is by communicating between government officials instead [@ackrenExploringRoleRegional2019; @harderArcticIndigenousRepresentation2024, 130].

The inclusion of anything related to the Greenlandic and Faroese representation within the Danish Realm into analyses of the political system of Denmark is not trivial, not least due to the separate party systems and because there are no institutional links between the Northatlantic MPs and neither their national governments nor parliaments [@westDanishRealmArctic2025, 602]. It is likely in this context that the activities of the four Northatlantic MPs feature only in the margins of research on Danish politics. This can be said for research on Folketing and its inner workings specifically, but also for adjacent topics such as research on elections and coalition formations as summarised above.

With this in mind, I assume that the activities of the Greenlandic and Faroese members of Folketinget have mostly been regarded not as a relevant aspect of the political system of Denmark that is worth considering, illustrated by Miller labelling those politicians, i.e. elected members of a legislative assembly, as "non-political" [-@millerFriendsRivals1996, 72][^slesvig]. In conclusion, there has been very little research activity on MPs from Greenland and the Faroe Islands and their behavior in Folketinget compared to the importance that is repeatedly being attributed to their presence. This has, however, changed in recent years with publications explicitly targeting Northatlantic representation in Folketinget which I am going to focus on in the following paragraphs. 

[^slesvig]: Miller, referring to the results of the 1957 election, also includes a representative for the German-speaking minority in Southern Slesvig in this group of non-political MPs. However, in this paper I will concentrate on Greenlandic and Faroese MPs only, not considering the representation of other minorities in Folketinget.

@harderFaeroskeOgGronlandske2022 

### generelt om aktivitet

"En yderligere lighed er, at både færøske og grønlandske medlemmer hovedsageligt engagerer sig i de sagsområder, der berører hjemlandet direkte" [@harderFaeroskeOgGronlandske2022, 20]

### spørgsmål

"parlamentariske spørgsmål (både udvalgsog § 20-spørgsmål), som medlemmerne stiller i perioden 2005-2019" [@harderFaeroskeOgGronlandske2022, 9]

"I perioden 2005-2019 stiller de grønlandske medlemmer næsten dobbelt så
mange parlamentariske spørgsmål som de færøske medlemmer (65 mod 35
pct.)" [@harderFaeroskeOgGronlandske2022, 10]

"Umiddelbart ser hverken det forhold, at et nordatlantisk medlem er fra et parti, der har rege- [SEITENUMBRUCH] ringsmagten hjemme, eller er fra en partifamilie, der har regeringsmagten i Danmark, ud til at påvirke medlemmets spørgeaktivitet signifikant" [@harderFaeroskeOgGronlandske2022, 10-11]

"Ligeså ser antallet af spørgsmål, som et færøsk medlem stiller, heller ikke ud til at hænge sammen med, om vedkommende repræsenterer et parti, der støtter rigsfællesskabet, eller et parti, der er imod dette" [@harderFaeroskeOgGronlandske2022, 11]

"I tråd med den tidligere litteraturs konklusioner viser den kvantitative indholdsanalyse af de parlamentariske spørgsmål, at de nordatlantiske mandater i perioden 2005-2019 hovedsageligt stiller spørgsmål af specifik relevans for Færøerne eller Grønland" [@harderFaeroskeOgGronlandske2022, 13]

"en større andel af både de færøske og grønlandske spørgsmål angår enten færinger eller grønlænderes forhold i Danmark eller oplysning om Færøerne eller Grønland i Danmark" [@harderFaeroskeOgGronlandske2022, 15]

"Generelt kan det således konkluderes, at de nordatlantiske mandater overordnet set har anvendt Folketingets spørgefunktion til at få information om specifikke forhold og indgå i kritisk dialog om specifikke problemstillinger, snarere end til at føre diskussioner om det overordnede eller fremtidige forhold mellem medlemmerne af rigsfællesskabet" [@harderFaeroskeOgGronlandske2022, 15]

### udvalgsarbejde

"Også de øvrige udvalgsmedlemskaber kan især forklares med, at man har
prioriteret at sidde i udvalg, der behandler områder af indenrigspolitisk betydning, som ikke er hjemtaget" [@harderFaeroskeOgGronlandske2022, 16]

"de grønlandske medlemmer har i langt højere grad været medlem af disse udvalg" [@harderFaeroskeOgGronlandske2022, 11]

### flertalsdannelse

"For så vidt angår den måde, hvorpå de nordatlantiske mandater i perioden 1998-2019 er indgået i de parlamentariske flertal, er den umiddelbare konklusion meget lig den, som Skjæveland finder for perioden 1953-1998: Når man har kunnet, har man påvirket tingenes tilstand. Desuden har man – ikke overraskende – overordnet set støttet regeringer, hvis partifarve man har delt" [@harderFaeroskeOgGronlandske2022, 18]

"Debatten er i særlig grad opstået i forbindelse med situationer, hvor det har været sandsynligt, at mandaterne kunne afgøre det parlamentariske flertal" [5]

"På samme måde har man forholdt sig forholdsvis ens til dannelsen af parlamentariske flertal: Nordatlantiske medlemmer har generelt ”spillet deres parlamentariske chance”, når den er opstået." [@harderFaeroskeOgGronlandske2022, 20]


@harderFaeroskeOgGronlandske2022
@ackrenExploringRoleRegional2019

@harderCrossparliamentaryControl2025
@harderCaseInnovativeParliamentary2022
@westPoliticalPartiesParliamentary2025

@cleggParliamentaryRepresentationOverseas2022
@harderArcticIndigenousRepresentation2024
@westDanishRealmArctic2025

Folketing is a strong institution for controlling the government @helboepedersenParliamentFOLKETINGETPowerful2020

 - apart from anecdotes [@millerFriendsRivals1996 about Mikael Gam 1960]: behavior "ved vi kun meget lidt om" [@harderFaeroskeOgGronlandske2022, 5; @harderCrossparliamentaryControl2025, 10]. (cf. contribution "with new knowledge concerning the dynamics between the Greenlandic representation in the Danish parliament and the Greenlandic government." [@harderCrossparliamentaryControl2025, 9])

when is it discussed? when is there relevance within DK? -> link to block on research on Rigsfællesskabet

  <!-- section Rigsfællesskabet -->

The recent increase in research output regarding the Northatlantic representation in Folketinget has to be seen in the context of the changed status of the Arctic within the Danish Realm and beyond, leading to a pressure to open the "'black box' of the Danish state" [@westDanishRealmArctic2025, 589] that the association of Denmark, Greenland and the Faroe Islands as *Rigsfællesskabet* and its dynamics were for a long time. Denmarks "Arctic awakening" [@jacobsenDanmarksArktiskeUdenrigspolitik2021] in the beginning of the 21^st^ century led to a shift from focussing on Danish-Greenlandic relations as domestic affairs within the Danish Realm to using them to legitimize Denmark as a geopolitical player in the Arctic instead.

This new geopolitical status of the Greenlandic and Faroese parts of *Rigsfællesskabet* is accompanied by an increased awareness of the colonialist heritage of Denmark in both Greenland and, to a lesser extent, the Faroe Islands [cf. @jensenPostcolonialDenmarkRot2015; @dyrendomgraugaardColonialReproductiveCoercion2025; @kociGreenlandFaroeIslands2021]. However, national parliaments are not in the spotlight of research dealing with Arctic Indigenous representation [@harderArcticIndigenousRepresentation2024, 130] which, in this case, would mainly concern MPs from Greenland, and not the Faroe Islands anyway.


- national parliaments not in focus of research on "Arctic indigenous peoples' political mobilisation" [@harderArcticIndigenousRepresentation2024, 130]
- beträfe nur GL, nicht FO

  + "et empirisk hul i rigsfællesskabets historie" [@harderFaeroskeOgGronlandske2022, 6]

link from Rigsfællesskabet to global postcolonialism discourse via overseas representation:

### Overseas Representation, Regional Parties, ethnic minorities?

### Cross-Parliamentary Control?

"I komparativt perspektiv er Folketingets udvalg i dag særligt væsentlige i forbindelse med arbejdet med at kontrollere regeringen (Holli og Harder, 2016). De nordatlantiske medlemmers udvalgsmedlemskab kan derfor – særligt i den senere del af perioden – være med til at give et indtryk af medlemmernes deltagelse i dette kontrolarbejde." [@harderFaeroskeOgGronlandske2022, 9]

## Research Gap

* Forschung zum Folketing klammert die vier nordatlantischen Mandate idR aus, daher sehr wenig Informationen zu ihrem Verhalten/ihrer pol. Arbeit. Ihre Relevanz wird vor allem im Zshg. mit Mehrheitsbeschaffungen attestiert, aber häufig nur anekdotisch bzw. letzten Endes als Kuriosität des politischen Systems in Dänemark behandelt
- link from @harderFaeroskeOgGronlandske2022; @harderCrossparliamentaryControl2025; @harderArcticIndigenousRepresentation2024 and @westPoliticalPartiesParliamentary2025 etc.: new data that has not been considered so far
* Mit meinem Datensatz kann ich die wenigen existierenden Papers, die die Arbeit dieser vier Abgeordneten systematisch bearbeiten, um den Aspekt Abstimmungsverhalten ergänzen.

specifically: GL/FO contributions, more generically: cross-parliamentary control, ethnic representation/Regional parties

"Desuden er de parlamentariske spørgsmål en interessant variabel, fordi de gi-
ver et indblik i, hvilke problematikker medlemmerne selv har valgt at sætte på
dagsordenen" [@harderFaeroskeOgGronlandske2022, 9] > Argumentation für/gegen roll call data?

#### use of roll call votes as method

# Dataset

I am using the *nordatlantisk-ft* dataset [@twenteNordatlantiskft2026] I compiled for this project, containing voting records of Folketinget members from Greenland and the Faroe Islands. Since Folketinget does not provide ready-to-use open datasets but only metadata [@berntzenParliamentaryOpenData2019, 9], I had to create the pipeline myself. The dataset is the result of a workflow in R, taking a list of MPs as input and then scraping voting records from Folketinget's Open Data Platform [@folketingetFolketingetsAbneData2026]. Both workflow and dataset are version-controlled using git, including long-term archival on Zenodo, and are accompanied by extensive metadata and documentation to ensure reproducibility [@mahrOneTemplateRule2025a]. [Codebook](https://mtwente.github.io/nordatlantisk-ft/docs/codebook.html) and [source code](https://github.com/mtwente/nordatlantisk-ft) are available via GitHub. 

For my analysis, I am going to use the data available as of `r format(file.info(here("data", "processed", "northatlantic_ft.rds"))$mtime, "%B %d, %Y")`. *nordatlantisk-ft* comprises `r length(unique(northatlantic_ft$vote_id))` records of votes cast by Northatlantic MPs between `r format(min(northatlantic_ft$date), "%B %d, %Y")` and `r format(max(northatlantic_ft$date), "%B %d, %Y")`. The dataset also contains metadata on each record -- e.g. date and topic of each vote or the overall roll call result, covering votes on `r length(unique(northatlantic_ft$roll_call_id))` proposals in Folketinget.

## Descriptive Overview

### MPs

Since 1953, two MPs are elected in both Greenland and the Faroe Islands to join the 175 MPs from Denmark proper in Folketinget. Before the 1953 reform, Greenland did not have any parliamentary representation in Folketinget, as opposed to the Faroe Islands with one guaranteed seat since 1849 [@harderFaeroskeOgGronlandske2022, 8-9; @kjaerDanskeFolketingsmedlemmerParlamentarisk2004]. According to @harderFaeroskeOgGronlandske2022 [7], it is usually experienced politicians that are elected to represent the Faroe Islands constituency, even though the job apparently has low priority compared to holding office 'at home' [@johannesenUniHolmJohannesen2016][^study].

[^study]: No comparable study exists for MPs from Greenland apart from the short MP profiles available in Danmarks Nationalleksikon [@billeFaeroskeFolketingsmedlemmer2025; @billeGronlandskeFolketingsmedlemmer2025].

In the time frame covered by Folketinget's Open Data Platform, there have been `r nrow(mp_names)` individual politicians from Greenland or the Faroe Islands who served as MPs (@tbl-MPs). `r sum(mp_names$origin == "GL")` of them are members of Greenlandic parties, and `r sum(mp_names$origin == "FO")` are representing Faroese parties. For MPs representing Greenland resp. the Faroe Islands in election periods earlier than `r format(min(northatlantic_ft$date), "%B %Y")`, see the list assembled by @harderSupplerendeMaterialeFaeroske2022.

`r sum(mp_names$substitute == TRUE)` of the politicians that are part of this dataset joined Folketinget in the role of a substitute member[^subst], filling in for one of the `r sum(mp_names$substitute == FALSE)` originally elected MPs while they were on leave [cf. @harderFaeroskeOgGronlandske2022, 12-13]. @fig-mandates shows membership periods and substitute members on a timeline for the duration that is covered by this dataset. Earlier terms served by Edmund Joensen and Lisbeth Petersen are cut off from the timeline since there are no voting records available online earlier than `r format(min(northatlantic_ft$date), "%B %d, %Y")`. 

[^subst]: Doris Jakobsen (SIU), Lisbeth Petersen (B), Magni Arge (E) and Sjúrður Skaale (C) all first became MPs as substitute members, but later joined as regular MPs after a general election.

```{r}
#| echo: false
#| message: false
#| label: fig-mandates
#| fig-height: 5
#| fig-cap: Timeline of MPs from Greenland and the Faroe Islands during the timeframe covered by this dataset. Substitutions shorter than three weeks are plotted as dots for better visibility.

# remove pre-2000 MPs (Edmund Joensen's and Lisbeth Petersen's first terms)
MP_dates <- MP_dates %>%
  filter(end >= as.Date("2000-01-01")) %>%
  mutate(origin = factor(origin, levels = c("GL", "FO")))

# Order MPs by first start date for nicer plotting
MP_dates <- MP_dates %>%
  group_by(origin, plotting_name) %>%
  mutate(first_start = min(start)) %>%
  ungroup()

# build facet-specific ordering
levels_gl <- MP_dates %>%
  filter(origin == "GL") %>%
  distinct(plotting_name, first_start) %>%
  arrange(desc(first_start)) %>%
  pull(plotting_name)

levels_fo <- MP_dates %>%
  filter(origin == "FO") %>%
  distinct(plotting_name, first_start) %>%
  arrange(desc(first_start)) %>%
  pull(plotting_name)

MP_dates <- MP_dates %>%
  mutate(
    plotting_name = case_when(
      origin == "GL" ~ factor(plotting_name, levels = levels_gl),
      origin == "FO" ~ factor(plotting_name, levels = levels_fo)
    )
  )

ggplot(MP_dates, aes(x = start,
                     xend = end,
                     y = plotting_name,
                     yend = plotting_name,
                     color = substitute)) +

  # for correct order of MPs across facets and scales
  geom_blank(aes(y = plotting_name)) +

  # Segments for all regular MPs and long-term substitutions
  geom_segment(
    data = MP_dates %>%
      filter(
        substitute == FALSE |
          (substitute == TRUE & !is.na(end) & as.numeric(end - start) >= 21)
      ),
    linewidth = 4,
    lineend = "butt"
  ) +

  # Points for short-term substitutions
geom_point(
  data = MP_dates %>%
    filter(
      substitute == TRUE &
        !is.na(end) &
        as.numeric(end - start) < 21
    ),
  aes(
    x = start,
    fill = substitute
  ),
  shape = 21,
  size = 1.5,
  stroke = 0.2,
  color = "#777777"
  ) +

  scale_color_manual(
    values = c("FALSE" = "#6195CF", "TRUE" = "#F7CB45"),
    labels = c("Regular", "Substitute")
  ) +
  scale_fill_manual(
  values = c("FALSE" = "#6195CF", "TRUE" = "#F7CB45"),
  guide = "none"
  ) +
  
  labs(color = "Membership") +

  # Folketing elections
  geom_vline(
    data = elections_ft,
    aes(xintercept = as.Date(date), linetype = election_type),
    color = "#000000",
    linewidth = 0.2
  ) +
  scale_linetype_manual(
    values = c("Folketing Election" = "longdash"),
    name = ""
  ) +

  facet_grid(
    origin ~ .,
    scales = "free_y",
    space = "free_y",
    labeller = labeller(origin = c("GL" = "Greenland", "FO" = "Faroe Islands"))
  ) +

  coord_cartesian(xlim = c(min(northatlantic_ft$date), NA)) +

  theme_nordatlantisk_basis() +
  theme(
    legend.position = "bottom",
    panel.grid.major.y = element_line(color = "#777777", linewidth = 0.05),
    panel.grid.minor.x = element_blank()
  )
```

### Political Parties

Since 2004, MPs from the Faroe Islands were members of `r n2w(sum(political_parties$origin == "FO"))` different political parties. Voters in Greenland elected politicians from `r n2w(sum(political_parties$origin == "GL") - 2)` different political parties. However, two of the Greenlandic representatives -- Aleqa Hammond and Aki-Matilda Høegh-Dam -- switched party affiliation during their time as member of Folketinget and joined two relatively young parties (Nunatta Qitornai and Naleraq[^3]).

[^3]: Both parties have been formed quite recently -- Naleraq in 2014 and Nunatta Qitornai in 2017. See @ahlnessNunattaQitornaiParty2020 for general information on Nunatta Qitornai and their policies. Nunatta Qitornai has de facto been dissolved after the 2021 Inatsisartut election [@hviidVittusQujaukitsoqMelder2025].

The Northatlantic MPs usually are members of the two biggest political parties of Greenland resp. the Faroe Islands [@elklitElectoralSystemFair2020]. These parties then act as de-facto ethnic regional parties in Folketinget [@ackrenExploringRoleRegional2019]. Before political parties were established in Greenland in the 1970s [@billeGronlandskeFolketingsmedlemmer2025; @michelsenGronlandPartisystemUdvikling1979], MPs were elected independently. Since then, one of the two Greenlandic seats has been held by social democratic *Siumut* (SIU) for most of the time. Only recently has Siumut not been represented in Folketinget, due to the MPs in question switching party affiliations (see @fig-party-timeline). The second Greenlandic MP was affiliated with centre-right *Atassut* (A) from 1977 to 2001. Since then, the MPs on the 'second' Greenlandic seat have been members of socialist *Inuit Ataqatigiit* (IA).

This dominance by two political parties is in contrast to the Faroese situation, where the four biggest parties take turns more often [@billeFaeroskeFolketingsmedlemmer2025]. In the time frame covered by this dataset, the Faroe Islands have been represented by one left-wing and one right-wing political party in Folketinget for most of the time, as opposed to the Greenlandic MPs having belonged to two left-wing parties.



todo: partigruppe 2001--2011: "Fra 2001 til 2011 dannede Siumuts og IA’s folketingsmedlemmer sammen med den færøske republikaner en nordatlantisk partigruppe." [@gadGronlandskeFolketingsmedlemmer2022], see chart by @harderSupplerendeMaterialeFaeroske2022 [S6]

"I 2001 sker der imidlertid det, at tre af de nordatlantiske mandater – de to grønlandske mandater samt medlemmer fra selvstændighedspartiet TF (T) – går sammen og danner Den Nordatlantiske gruppe, der med intern koor- dinering, arbejdsdeling og fælles sekretariat minder om de danske partigrupper. Denne gruppe består indtil valget i 2011, om end Lars Emil Johansen (SIU) melder sig ud af gruppen allerede i 2009" [@harderFaeroskeOgGronlandske2022]

"Fra 2007 sker der i øvrigt den ændring, at nordatlantiske medlemmer ikke længere meddeler deres medlemskab af partigrupper til Folketingets administration og derfor ikke er officielt registreret som sådan i Folketingets dokumenter efter 2007. Imidlertid er flere af de nordatlantiske medlemmer – også efter 2007 – i praksis medlem af danske partigrupper" [@harderFaeroskeOgGronlandske2022, 17]

All represented political parties are listed in the [Appendix](#list-of-parties) with additional information (where available) on their positions from the ParlGov dataset [@doringParlGov2024Release2024]. @fig-party-timeline shows the allocation of Folketinget seats to political parties from Greenland and the Faroe Islands on a timeline.

```{r}
#| echo: false
#| message: false
#| label: fig-party-timeline
#| fig-cap: Timeline of political parties from Greenland and the Faroe Islands with an MP in Folketinget since 2004. Dashed lines indicate election dates.

# collapse MP_dates timeline into non-overlapping party intervals
party_periods <- MP_dates %>%
  select(party, origin, start, end) %>%
  arrange(party, origin, start, end) %>%
  group_by(party, origin)

# join colours from political_parties
party_periods <- party_periods %>%
  left_join(political_parties %>% select("party", "full_name", "colour"), by = "party")

# sort parties with start date
party_periods <- party_periods %>%
  group_by(origin) %>%
  mutate(party = factor(party, levels = rev(unique(party[order(start)])))) %>%
  ungroup()

# set party name labels including UF
y_labels <- setNames(party_periods$full_name, party_periods$party)
y_labels <- c(y_labels, "UF" = "No Party Affiliation")

# set UF colour to grey
party_periods$colour[party_periods$party == "UF"] <- "#777777"

ggplot(party_periods, aes(x = start, xend = end, y = party, yend = party, color = colour)) +
  geom_segment(linewidth = 8, lineend = "butt") +
  scale_color_identity() +

  scale_y_discrete(labels = y_labels) +
  
  # add Folketing Election dates
  geom_vline(data = elections_ft, aes(xintercept = as.Date(date), linetype = election_type),
             color = "#000000", linewidth = 0.2) +
  scale_linetype_manual(values = c("Folketing Election" = "longdash"), name = "") +
  
  facet_grid(origin ~ ., scales = "free_y", space = "free_y",
             labeller = labeller(origin = c("FO" = "Faroe Islands", "GL" = "Greenland"))) +
  
  coord_cartesian(xlim = c(min(northatlantic_ft$date), NA)) +
  
  theme_nordatlantisk_basis() +
  theme(
    legend.position = "none",
    panel.grid.major.y = element_line(color = "#777777", linewidth = 0.05),
    panel.grid.minor.x = element_blank()
  )
```

### Voting Records

There are 179 MPs in Folketinget, the parliament of Denmark. But only 175 of them are elected in Denmark proper, while the remaining four seats are guaranteed for the so-called North Atlantic MPs from Greenland and the Faroe Islands, which both are autonomous members of the Danish Realm today after being under Danish colonial rule for centuries. Their presence in Folketinget can prove important for domestic Danish politics because not all of the four MPs join one of the two party blocs and can thus provide decisive votes for passing legislation and even for securing the national government's stability [@harderFaeroskeOgGronlandske2022, 19]: Several times, both Danish party blocs found themselves with election results where they did not have what is called a "domestic majority" [_indenrigspolitisk flertal_, @buggeRegeringensIndenrigspolitiskeFlertal2023] but instead relied on the North Atlantic MPs to form a majority of 90 seats -- using the Danish expression: to be able to count to 90. Recently, this was the case in 2016 and 2022[^1] [@sorensenEfterBrokBla2022]: In both instances, opposition parties cried foul and called for re-evaluation of the North Atlantic MPs' roles, who have questioned these arrangements themselves as well [e.g. @skaaleFaeroskMFerNordatlantiske2022; @skaale20sporgsmalUS182021; @joensenLagtingsformandenVilAfskaffe2019], even though they have profited from them -- e.g. by being awarded chairmanship of parliamentary commissions [@harderFaeroskeOgGronlandske2022, 13]. <!-- 2023 situation mit moderaterne hinzufügen! -->

- data sources
- creation process
- data model
- description

```{r}
#| echo: false
#| message: false
#| fig-height: 1.2
#| fig-env: "figure*"
#| label: fig-absence-share
#| fig-cap: Northatlantic MPs are listed as absent in 98% of the observations in this dataset.

# Compute proportions of vote types
vote_props <- northatlantic_ft %>%
  count(vote_type_id) %>%
  mutate(
    share = n / sum(n),
    vote_type = factor(vote_type_id,
                       levels = c(3, 4, 2, 1),
                       labels = c("absent", "abstention", "against", "for"))
  )

# Single horizontal bar plot
ggplot(vote_props, aes(x = 1, y = share, fill = vote_type)) +
  geom_col(width = 0.5) +
  coord_flip() +
  scale_fill_manual(
    name = "Vote",
    labels = c("for", "against", "abstention", "absent"),
    values = c("#6195CF", "#F7CB45", "#777777", "#D1BBD7"),
    breaks = c("for", "against", "abstention", "absent")
  ) +

  scale_y_continuous(
    labels = scales::label_percent(),
    breaks = c(0, 0.05, 0.25, 0.50, 0.75, 1.00),
    expand = expansion(mult = c(0, .015))
  ) +
  
  labs(x = NULL, y = NULL) +
  theme_nordatlantisk_basis() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.ticks.x = element_line(color = "#000000", linewidth = 0.1),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.position = "top"
  )
```

# Research Design

## Theoretical Approach

<!-- - case selection/overview data (temporal and spatial), unit of analysis
- measurement dependent variable (and data source)
- measurement independent and control variables (and data source)
- methodological approach (why this one?) -->

  - if they have power, they use it [@harderFaeroskeOgGronlandske2022, 18; @skjaevelandModellingGovernmentFormation2009, 723]
  + relevans efter valg 1971 [@gadGronlandskeFolketingsmedlemmer2022; @kurrild-klitgaardElectionInversionsCoalitions2013, 131; @skjaevelandModellingGovernmentFormation2009, 723], 2007 om asylcentre, but was resolved without GL/FO influence [@gadGronlandskeFolketingsmedlemmer2022; @cleggParliamentaryRepresentationOverseas2022, 239], relevans efter valg 2007, 2011, 2015 [@harderFaeroskeOgGronlandske2022, 19] and 1998 [@skjaevelandModellingGovernmentFormation2009, 723], and 1960 [@skjaevelandGovernmentFormationDenmark2003, 448], and 1957 [@millerFriendsRivals1996, 71]. 2015 not relevant after all? [@arndtFolketingswahlDanemarkVom2019, 785]
  - "I to tilfælde har grønlandsk valgte MFere opnået ministerposten" [@michelsenGronlandPartisystemUdvikling1979, 55] > Gam og hvem?
  - @dallGronlandOgFaeroerne2022 with comprehensive overview

  - "Den anden er Edmund Joensen (SP), der i perioden lige efter valget i 2007 spillede en afgørende rolle i at sikre den borgerlige regerings parlamentariske flertal. Som det eneste færøske medlem i hele perioden blev han fuldt medlem af hele fem folketingsudvalg samt suppleant i et enkelt. Noget tyder således på, at det at sikre regeringens flertal kan blive afspejlet i blandt andet antallet af udvalgsposter, som man råder over – en antagelse, der i øvrigt underbygges af, at Aleqa Hammond (SIU/NQ) blev tildelt den symbolsk vigtige rolle som formand for Grønlandsudvalget efter at have indgået aftale om at støtte den borgerlige regering" [@harderFaeroskeOgGronlandske2022, 17]
  - "Analysen ovenfor viser imidlertid, at der ikke er nogen signifikant sammenhæng mellem antallet af parlamentariske spørgsmål, som et nordatlantisk medlem stiller, og spørgsmålet om, hvorvidt vedkommendes partifamilie støttede regering eller opposition" [@harderFaeroskeOgGronlandske2022, 17]

"Overordnet set kan det fastslås, at de grønlandske medlemmer nu som før del-
tager langt mere aktivt i folketingsarbejdet, end de færøske medlemmer gør" [@harderFaeroskeOgGronlandske2022, 19]

### Dataset Overview

## Hypotheses

<!-- * Hypothesis: causal relationship between two variables: the more, the less...
* Hypothesis in its own paragraph and in italics -->

## Empirical Strategy

<!-- - variables and controls
- unit of analysis
- method (logit model etc.) -->

- use tredje behandling? cf. @green-pedersenBlokpolitikOgNye2020 [315]

"Fordi flere af de nordatlantiske medlemmer holder længere orlov fra Folketinget, er data herfor indhentet på parti- og ikke individniveau" [@harderFaeroskeOgGronlandske2022, 9]

# Results

<!-- - per hypothesis -->

## Implications, Future Research

## Critical Assessment

<!-- * intro&conclusion should speak to each other/mirror each other -->

### Introduction

When the government of Denmark only has a narrow majority in parliament, four MPs regularly come into the spotlight: The so-called North Atlantic MPs -- two from Greenland, two from the Faroe Islands -- who usually abstain from voting on matters not concerning Greenland or the Faroe Islands but might be able provide four sometimes crucial votes to form a majority. In this paper, I am going to investigate what role these four MPs tend to play in Folketing: When do they capitalise on their voting power? Do they actually have as much power as the numbers suggest? Or is their influence rather a media frenzy, highlighting a rather unusual trait of day-to-day domestic politics?

There are 179 MPs in Folketinget, the parliament of Denmark. But only 175 of them are elected in Denmark proper, while the remaining four seats are guaranteed for the so-called North Atlantic MPs from Greenland and the Faroe Islands, which both are autonomous members of the Danish Realm today after being under Danish colonial rule for centuries. Their presence in Folketinget can prove important for domestic Danish politics because not all of the four MPs join one of the two party blocs and can thus provide decisive votes for passing legislation and even for securing the national government's stability [@harderFaeroskeOgGronlandske2022, 19]: Several times, both Danish party blocs found themselves with election results where they did not have what is called a "domestic majority" [_indenrigspolitisk flertal_, @buggeRegeringensIndenrigspolitiskeFlertal2023] but instead relied on the North Atlantic MPs to form a majority of 90 seats -- using the Danish expression: to be able to count to 90. Recently, this was the case in 2016 and 2022[^1] [@sorensenEfterBrokBla2022]: In both instances, opposition parties cried foul and called for re-evaluation of the North Atlantic MPs' roles, who have questioned these arrangements themselves as well [e.g. @skaaleFaeroskMFerNordatlantiske2022; @skaale20sporgsmalUS182021; @joensenLagtingsformandenVilAfskaffe2019], even though they have profited from them -- e.g. by being awarded chairmanship of parliamentary commissions [@harderFaeroskeOgGronlandske2022, 13]. <!-- 2023 situation mit moderaterne hinzufügen! -->

[^1]: Even though the North Atlantic MPs' role was discussed extensively in 2022, it was resolved without the North Atlantic MPs by forming an unusual cross-bloc coalition government.

Just as political relations between Denmark, Greenland and the Faroe Islands are being scrutinised time and again in terms of economy, ecology, security and last but not least postcolonialism, so is the position of two MPs elected to Folketinget in Greenland and the Faroe Islands, respectively.

* In general, when do the Northatlantic MPs take part in roll call votes? Hypothesis: The MPs are more likely to take part in voting when the issue concerns Greenland/policy areas that GL/FO do not have authority over (e.g. Defense, based on keywords in proposal description and the ministerområde responsible for the proposal, e.g. Forsvarsministeriet, Finansministeriet etc.)
* The 2x2 MPs are conceived as one group by DK politicians and media. Do they also vote like one group? Hypothesis: After excluding absences, similar voting behavior can not be observed across Greenlandic and Faroese MPs, but across the political blocks the MPs belong to (left/right -- in the dataset: usually 2x left from GL, and 1x each from FO, but never multiple MPs from the same party at the same time)
* When Northatlantic MPs have the chance to have influence, they act on it (based on decades of more or less anecdotal evidence). So, does the seat share of Danish coalition parties influence voting behaviour of Northatlantic MPs? Hypothesis: The fewer FT mandates the Danish cabinet coalition parties have themselves, the more likely are GL/FO MPs to cast a vote (regardless of the content of the proposal being voted on)
  * Based on questioning/committee activity analysis of Northatlantic MPs: Cross-parliamentary control mechanisms are especially made use of if the MP's own party is an opposition party in GL/FO. Transferring this finding onto another type of parliamentary activity, can the Northatlantic MP's voting behavior also be considered a form of cross-parliamentary control? Hypothesis: MPs are more likely to engage in parliamentary activity in FT, in this case taking part in voting, if their own party is in opposition in GL/FO.

# References {.unnumbered}

::: {#refs}
:::

\onecolumn

# Appendix {.appendix}

## List of MPs {.appendix .unnumbered .unlisted}

```{r}
#| echo: false
#| message: false
#| label: tbl-MPs
#| tbl-colwidths: [12, 30, 25, 10, 10, 13]

MPs <- mp_names %>%
  select(-wikidata_id) %>%
  select(origin, everything()) %>%
  arrange(origin, surname)

printable_names <- c("Origin", "Surname(s)", "First Name(s)", "MP ID", "Party", "Substitute")

kable(MPs, caption = "List of MPs from Greenland and the Faroe Islands in Folketinget (October 2004 -- January 2026)", col.names = printable_names)
```

## List of Parties {.appendix .unnumbered .unlisted}

Information on each party's positions are taken from the 2024 ParlGov release [@doringParlGov2024Release2024]. ParlGov does, however, not include metadata on the two Greenlandic parties that have been represented in Folketinget only recently -- Nunatta Qitornai in 2018/2019 and, since 2025, Naleraq. Additionally, Aleqa Hammond was MP without party affiliation from `r format(MP_dates$start[MP_dates$period_id == "hammond_2"], "%B %d, %Y")` to `r format(MP_dates$end[MP_dates$period_id == "hammond_2"], "%B %d, %Y")` [@twenteNordatlantiskft2026].

```{r}
#| echo: false
#| message: false
#| label: tbl-parties-gl
#| tbl-colwidths: [30, 30, 10, 10, 10, 10]

parties_gl <- political_parties |>
  filter(origin == "GL") |>
  mutate(
    full_name_abbr = paste0(full_name, " (", party, ")")
  ) |>
  select(
    full_name_abbr, party_family,
    left_right, state_market,
    liberty_authority, anti_pro_eu
  )

parties_fo <- political_parties |>
  filter(origin == "FO") |>
  mutate(
    full_name_abbr = paste0(full_name, " (", party, ")")
  ) |>
  select(
    full_name_abbr, party_family,
    left_right, state_market,
    liberty_authority, anti_pro_eu
  )

printable_names <- c(
  "Party", "Party Family",
  "Left–Right", "State–Market", "Liberty–Authority", "Anti–Pro-EU"
)

kable(
  parties_gl,
  caption = "List of political parties from Greenland with an MP in Folketinget since 2004",
  col.names = printable_names,
  row.names = FALSE,
  escape = FALSE
)
```

```{r}
#| echo: false
#| message: false
#| label: tbl-parties-fo
#| tbl-colwidths: [30, 30, 10, 10, 10, 10]

kable(
  parties_fo,
  caption = "List of political parties from the Faroe Islands with an MP in Folketinget since 2004",
  col.names = printable_names,
  row.names = FALSE,
  escape = FALSE)
```

## Word Count {.appendix .unnumbered .unlisted}

Currently featuring {{wordcount}} words without references. Including references, there are {{wordcountref}} words in this document.

\twocolumn
